### 2024-04-18 paper/main.tex
Reimport from the "corrupted repo"

% chktex-file 18 
% chktex-file 46

\documentclass{isprs} % isprs class modified 23-04-2019 (Dennis Wittich)
\usepackage{subfigure}
\usepackage{setspace}
\usepackage{geometry} % added 27-02-2014 Markus Englich
\usepackage{epstopdf}
\usepackage[labelsep=period]{caption}  % added 14-04-2016 Markus Englich - Recommendation by Sebastian Brocks
\usepackage[british]{babel} 
\usepackage[hang]{footmisc}
\def\footnotemargin{1em} % added 08-01-2020 Dennis Wittich

% //-- HL:
\usepackage{booktabs}
% \usepackage{flushend}
\usepackage{siunitx}
\usepackage{threeparttable}
\usepackage{url}
\newcommand{\ie}{ie}
\newcommand{\eg}{eg}
\usepackage{listings}
\usepackage{color}
\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}
\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},   
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\footnotesize,
    breakatwhitespace=false,         
    breaklines=true,                 
    captionpos=b,                    
    keepspaces=true,                 
    numbers=left,                    
    numbersep=5pt,                  
    showspaces=false,                
    showstringspaces=false,
    showtabs=false,                  
    tabsize=2
}
\lstset{style=mystyle,basicstyle=\ttfamily\footnotesize,breaklines=true}
% --//

\usepackage[authoryear]{natbib}
\def\bibhang{0pt}

\geometry{a4paper, top=25mm, left=20mm, right=20mm, bottom=25mm, headsep=10mm, footskip=12mm} % added 27-02-2014 Markus Englich
%\usepackage{enumitem}

%\usepackage{isprs}
%\usepackage[perpage,para,symbol*]{footmisc}

%\renewcommand*{\thefootnote}{\fnsymbol{footnote}}
\captionsetup{justification=centering,font=normal} % thanks to Niclas Borlin 05-05-2016
\captionsetup[figure]{font=small} % added 23-04-2019 Dennis Wittich
\captionsetup[table]{font=small} % added 23-04-2019 Dennis Wittich

\begin{document}

\title{Streaming CityJSON datasets}

% KAO: Remove extra spacing
\author{
 Hugo Ledoux\textsuperscript{1}, Gina Stavropoulou\textsuperscript{1}, Balázs Dukai\textsuperscript{2}}

% KAO: Remove extra newline
\address{
	\textsuperscript{1 }Delft University of Technology, the Netherlands---\texttt{[h.ledoux, g.stavropoulou]@tudelft.nl}\\
	\textsuperscript{2 }3DGI, the Netherlands---\texttt{balazs.dukai@3dgi.nl}\\
}

% If the corresponding author is NOT the final author, always add a % space before the subsequent comma, i.e.
% first author name\textsuperscript{a,}\thanks{Corresponding author} , % second author name \textsuperscript{b}, etc.
% thanks to Niclas Borlin 05-05-2016
% information on the corresponding author should not be used any longer and has been commented out
% C. Heipke, Jan 03,2024

% the use of the information of commissions and working groups should not be used any longer and has been commented out
% C. Heipke, Sept. 20,2022
%\commission{XX, }{YY} %This field is optional. If filled, XX and YY should be replaced by adequate numbers. See https://www2.isprs.org/commissions/
%\workinggroup{XX/YY} %This field is optional.
%\icwg{}   %This field is optional.

\abstract{
% The abstract must have the following structure: 
% \begin{enumerate}
% \setlength\itemsep{0em}\setlength\parskip{0em}\setlength\topsep{0em}\setlength\partopsep{0em}\setlength\parsep{0em} 
% \item{Title of the contribution} 
% \item{Author(s) and affiliation}
% \item{Keywords (max. 6)}
% \item{Main body (one page max)}
% \item{Illustrations (one page max, optional)}
% \item{References (8 max, optional)}
% \end{enumerate}
% Abstracts must contain the title, the author(s) names incl. affiliations, the keywords and a summary of the contribution (research question, relevance, solution, evaluation).

We present \emph{CityJSON Text Sequences} (CityJSONSeq), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
CityJSONSeq was added to the CityJSON version 2.0 standard that was recently released, and it allows us to stream very large 3D city models.
The main idea is to decompose a CityJSON file into its individual features (eg each Building, each Bridge, etc.) and create several JSON objects of a newly defined type \texttt{CityJSONFeature}.
We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to/from CityJSONSeq, and discuss different aspects of the resulting CityJSONSeq, \eg\ size, usability, etc.

We think that CityJSONSeq is a better format than CityJSON
CityJSONSeq is at about 10\% more compacter, takes an order of magnitude less time to process and uses significantly less memory (for some operations, not all).

}

\keywords{3D city modelling, CityJSON, CityGML, streaming}
\maketitle
%\saythanks % added 28-02-2014 Markus Englich

\sloppy

%%%
%
\section{Introduction}%
\label{sec:intro}

CityJSON is a JSON-based encoding for storing 3D city models that implements a subset of the latest version, v3.0.0, of the CityGML data model~\citep{OGC-CityGML3}.
As further described in \citet{19_ogdss_cityjson}, its development started in 2017 with the aim of offering an easy-to-use and web-ready alternative to the GML-encoded CityGML files~\citep{OGC-CityGML3-XML}, which in practice can be rather verbose, difficult to parse, and complex to manipulate.
The first official release of CityJSON (version 1.0) was a success: 
\begin{enumerate}
  \item its JSON-based files were on average around 7X more compact than their CityGML-GML equivalents without loss of information. See \citet{19_ogdss_cityjson} and \url{https://www.cityjson.org/filesize/}, and also \citet{Praschl23} for a comparison with standard computer graphics formats; 
  \item it was adopted as an OGC Community Standard~\citep{OGC-CityJSON-v10}; 
  \item it was adopted by the Dutch government as a 3D standard to distribute nation-wide 3D datasets; 
  \item several implementations, \eg\ FME, were developed.
\end{enumerate}

%

However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries (see Figure~\ref{fig:cj_idea}) made the \emph{streaming} of very large datasets complex, if not impossible.
Because more and more datasets from cities are very large (it is not uncommon to have CityGML-GML files over 2GB), this made the processing of CityJSON difficult.
As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.

%

We present in this paper \emph{CityJSON Text Sequences} (\emph{CityJSONSeq} in short in the following), a format based on \emph{JSON Text Sequences}~\citep{IETF-JSONSeq} and CityJSON, and inspired by \emph{GeoJSON Text Sequences}~\citep{IETF-GeoJSONSeq}. 
CityJSONSeq was added to the CityJSON version 2.0 standard that was released in 2023 (and also standardised by the OGC, see \citet{OGC-CityJSON-v20}).
As further described in Section~\ref{sec:cityjsonseq}, the idea is to decompose a (often large) CityJSON file into its individual \emph{features} (\eg\ each \texttt{Building}, each \texttt{Bridge}, etc.) and create several JSON objects of a newly defined type \emph{CityJSONFeature}.
These objects are either serialised into a text file, or streamed from a server to client or from one application to another.
This allows us to avoid a long list of vertices that need to be indexed, as it is the case with CityJSON files (see Section~\ref{sec:cityjson} for details).
\begin{figure}
  \centering
  \includegraphics[width=0.95\linewidth]{figs/cj_idea}
  \caption{An example of a CityJSON file. The vertices are stored in a global list, and the position of the vertices in that list are used in the geometries (represented by the arrows, many have been left out for clarity).}%
\label{fig:cj_idea}
\end{figure}

%

We have implemented and released an open-source program to convert between in either direction between CityJSON and CityJSONSeq objects; we describe it in Section~\ref{sec:experiments}.
We also analyse the size of CityJSONSeq files versus CityJSON files, and that for several real-world datasets.
It can be observed that one advantage of CityJSONSeq, besides that files containing thousands of features can be processed with a low memory footprint, is that it compresses further the CityJSON files by around 12\%, sometimes more.
We discuss in Section~\ref{sec:experiments} the reasons for this interesting finding.
Finally, we will also present the open-source software that are currently being developed to process streams of CityJSON datasets.

%%%
%
\section{Structure of a CityJSON file}%
\label{sec:cityjson}

% \begin{itemize}
%   \item global indexing of vertices make streaming impossible
%   \item discuss the geometry templates and other metadata
% \end{itemize}

As shown in Figure~\ref{fig:cj_idea}, a CityJSON object, which is a JSON object, represents a given geographical area, and it typically contains the following JSON properties: 
\begin{enumerate}
  \item \texttt{"type"}: it must be \texttt{"CityJSON"};
  \item \texttt{"version"}: \texttt{"2.0"} is the current version;
  \item \texttt{"metadata"}: different metadata related to the dataset can be stored. The most important is the definition of the coordinate reference system (CRS)
  \item \texttt{"transform"}: CityJSON \texttt{"vertices"} are compressed and stored with integers only. The parameters of this property allows us to convert from those to real-world coordinates.
  \item \texttt{"CityObjects"}: a dictionary where the properties are the identifiers of the city objects (\emph{IDs}), which are any CityGML city objects (for instance a \texttt{Building}, a \texttt{BuildingPart}, a \texttt{SolitaryVegetationObject}, etc.).
  Each of the city objects are listed one after the others, even if some are \texttt{"children"} of others 
  As an example, for a \texttt{Building} containing 2 parts, the 3 objects will be represented at the same level and linked by their \emph{IDs}, as shown in Figure~\ref{fig:parents_children}. The schema is thus flat and all hierarchies have been removed.
  Each city object can have a \texttt{"parents"} and/or a \texttt{"children"} property, and this is how in the snippet the building \texttt{"id-1"} is linked to its 2 parts.
  % We define 2nd-level city objects as having a \texttt{"parent"}.
  The fact that a dictionary is used means that developers have direct access to the city objects through their IDs (and also in constant time if a hashmap is used to implement the dictionary while parsing the file).
  \item \texttt{"vertices"}: The 3D geometric primitives are those of CityGML, and multi/composite solids having several parts are supported.
  A geometric primitive does not list all the coordinates of its vertices, rather the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and geometric primitives refer to the position of a vertex in that array.
  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is reused, because it has been used for many years, with success, in the computer graphics community.
  There are several advantages to this approach.
  First, the files can be compressed: 3D vertices are often shared by several surfaces, and repeating them can be costly (especially if they are very precise, often sub-millimetre is used).
  Second, this increases the topological relationships that are explicitly stored in the file, and several operations (\eg\ are 2 buildings adjacent?) can be sped up and made more robust.
  Third, it is very easy to convert to a representation listing all coordinates; the inverse is not true. 
  However, this is where the streaming of geometries is problematic: the array of "vertices" can for very large areas, contains several millions vertices.
  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can be for streaming first waiting for millions of unused vertices to be transferred.
  \item \texttt{"appearances"}: Both textures and materials for surfaces are supported.  
  The material of a surface is represented with the X3D specifications\footnote{\url{https://en.wikipedia.org/wiki/X3D}}, and for the textures the COLLADA specifications\footnote{\url{https://www.khronos.org/collada/}} are reused.
  \item \texttt{"geometry-templates"} 
  Geometry templates are used to represent identical geometries several times, they only need to be defined once (and translations/rotations/scaling are applied). They are used for city objects like trees and lamp posts. 
\end{enumerate}

\begin{figure}
  \centering
\begin{lstlisting}
  "CityObjects": {
    "id-1": {
      "type": "Building",
      "attributes": {...},
      "children": ["id-2", "id-3"],
      "geometry": [{...}]
    },
    "id-2": {
      "type": "BuildingPart",
      "parents": ["id-1"],
      "geometry": [{...}]
      ...
    },
    "id-3": {
      "type": "BuildingPart",
      "parents": ["id-1"],
      "geometry": [{...}]
      ...
    }
    ...
    "id-77": {}
  }
\end{lstlisting}
  \caption{CityJSON mechanism to flatten out the schema: each city object are stored in a flat list, and they are linked together with the properties \texttt{"parents"} and \texttt{"children"}.}%
\label{fig:parents_children}
\end{figure}

%%%
%
\section{Streaming (3D) datasets}%
\label{sec:streaming}

In the context of geo-information, a \emph{stream} is a sequence of data that is available over a period of time, and ``can be thought of as items on a conveyor belt being processed one at a time rather than in large batches''\footnote{From \url{https://en.wikipedia.org/wiki/Stream_(computing)}}.

%

For GML-based formats (which are feature-centric), modifying a file for the purpose of streaming is usually a simple task that involves sending the features in the dataset one-by-one.
However, notice that this is only true for simple files and GML features stored following the Simple Feature paradigm~\citep{OGC-SF}.
For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used (for reusing geometries and for storing some topological relationships), the conversion often needs a large amount of pre-processing.
The presence of \emph{XLinks} means that references between elements are possible (\eg\ reusing surfaces), but if in the stream the referenced surface has not appeared, the receiver cannot process the information.
It is not always possible to order a CityGML files so that all the references can be resolved without having to store locally information until it appears in the stream.
It is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.

%

For formats that use a global indexing of vertices (as written above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
\citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that vertices are not to be used anymore (and thus can be freed from memory).
This allows us, in theory, to process/edit/manipulate infinitely large meshes, since they never have to be completely loaded in memory.
The idea is exemplified by the construction of gridded terrains that are gigabytes in size~\citep{Isenburg06-1}.

However, this cannot be implemented in CityJSON directly as all the vertices need to be in the one JSON property \texttt{"vertices"}.

%%%
%
\section{CityJSON Text Sequences}%
\label{sec:cityjsonseq}

\begin{figure*}[h]
  \centering
  \includegraphics[width=0.95\linewidth]{figs/cjseq_idea}
  \caption{The CityJSONSeq of a CityJSON dataset with two buildings contains three JSON objects: one for the metadata, plus one for each building.}%
\label{fig:cjseq_idea}
\end{figure*}

As shown in Figure~\ref{fig:cjseq_idea}, a CityJSONSeq decomposes a CityJSON object into its \emph{City Objects} (or ``features'', for instance each building, each bridge, each road, etc.), to create a sequence of several JSON objects.
Those JSON objects are of a new type that was introduced in CityJSON v2.0: \texttt{CityJSONFeature}.
It allows the storage of a single feature, for instance a \texttt{Building} together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}). 
Unlike a CityJSON Object, each feature is independent. 
It has its own list of vertices (which is thus \emph{local} to be JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
The allowed properties are shown below; notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are many children.
\begin{lstlisting}
{
  "type": "CityJSONFeature",
  "id": "id-1", 
  "CityObjects": {
    "id-1": {
      "type": "Building", 
      "attributes": { 
        "roofType": "gabled roof"
      },
      "children": ["mybalcony"],
      "geometry": [...]
    },
    "mybalcony": {
      "type": "BuildingInstallation", 
      "parents": ["id-1"],
      "geometry": [...]
    }
  },
  "appearance": {...}
  "vertices": [...]
}
\end{lstlisting}

%

CityJSONSeq follows the specifications of \texttt{ndjson} (newline delimited JSON)\footnote{\url{https://github.com/ndjson/ndjson-spec/}} and two constraints are added for handling CityJSON\@:
\begin{enumerate}
  \item each JSON Object must conform to the \emph{JSON Data Interchange Format specifications}~\citep{IETF-JSON} and be written as a UTF-8 string;
  \item each JSON Object must be followed by a new-line (LF: \texttt{"\textbackslash n"}) character, and it may be preceded by a carriage-return (CR: \texttt{"\textbackslash r"});
  \item a JSON Object must not contain the new-line or carriage-return characters;
  \item the first JSON Object must be of type \texttt{CityJSON};
  \item the following JSON Objects are of type \texttt{CityJSONFeature}.
\end{enumerate}

Note that a \texttt{CityJSONFeature} object does not contain all the information that is required for parsing the feature. 
Most commonly, the \texttt{"transform"} property, the CRS, and the \texttt{"geometry-templates"} must be known by the client in order to correctly reconstruct and georeference the city objects. 
The rule \#4 ensures that those are available. 
The CityJSON object must contain \texttt{"transform"} and eventually the other properties if needed; \texttt{"CityObjects"} and \texttt{"vertices"} must be present but be empty (to ensure that the JSON object is valid).

%

The CityJSONSeq for Figure~\ref{fig:cjseq_idea} is shown in Figure~\ref{fig:stream}:
\begin{figure*}
  \begin{lstlisting}
  {"type":"CityJSON","version":"2.0","transform": {"scale":[1.0,1.0,1.0],"translate":[0.0, 0.0, 0.0]},"metadata":{"referenceSystem":"https://www.opengis.net/def/crs/EPSG/0/7415"},"CityObjects":{},"vertices":[]}
  {"type":"CityJSONFeature","id":"id-1","CityObjects":{...},"vertices":[...]} 
  {"type":"CityJSONFeature","id":"id-2","CityObjects":{...},"vertices":[...]} 
  \end{lstlisting}
\caption{An example of a CityJSONSeq stream containing 3 features.}%
\label{fig:stream}
\end{figure*}

%%%
%
\section{Experiments with real-world datasets}%
\label{sec:experiments}

To convert efficiently between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, freely available at \url{https://github.com/cityjson/cjseq/}.
The program takes care of not only the geometries, but also of the materials, textures, and geometry templates that the dataset could contain.
It contains two sub-commands: 
\begin{enumerate}
  \item \texttt{cat}: based on a CityJSON file, it outputs a CityJSONSeq;
  \item \texttt{collect}: it reads a CityJSONSeq from \emph{stdin}, collects the features, and outputs a CityJSON file (duplicates vertices are merged).
\end{enumerate}

%%%
\subsection{File size comparison}

\begin{table*}
  \centering
  \begin{threeparttable}
  \caption{The datasets used for the benchmark. }%
  \label{tab:datasets}
  \small
  \begin{tabular}
    {@{}lcccccrrrcrr@{}}\toprule
    &&  \multicolumn{3}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{2}{c}{\textbf{vertices}}   \\ 
    \cmidrule{3-5} \cmidrule{7-9} \cmidrule{11-12} 
     && CityObjects &  LoDs & appearance\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ \\
    \midrule
    \textbf{3DBAG}          && \qty{1110} bldgs & 0+1.2+1.3+2.2 & &&    \qty{6.7}{\mega\byte} & \qty{4.3}{\mega\byte} &  64\%  &&   \num{82509} &  \num{4112} \\
    \textbf{Helsinki}       && \qty{77231} bldgs   & 1+2   &         && \qty{600}{\mega\byte} & \qty{432}{\mega\byte} &  72\%  && \num{3038576} &  \num{2202} \\
    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & 1+2   & tex     && \qty{748}{\mega\byte} & \qty{675}{\mega\byte} &  90\%  && \num{3038576} &  \num{2202} \\
    \textbf{Ingolstadt}     && \qty{55} bldgs      & 3     &         && \qty{5.1}{\mega\byte} & \qty{4.0}{\mega\byte} &  78\%  &&   \num{87972} & \num{12800} \\
    \textbf{Montréal}       && \qty{294} bldgs     & 2     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} &  85\%  &&   \num{31585} &  \num{3393} \\
    \textbf{NYC}            && \qty{23777} bldgs   & 2     &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} &  90\%  && \num{1035804} &  \num{2608} \\
    \textbf{Railway}        && \qty{50} diff       & 3     & tex+mat && \qty{4.5}{\mega\byte} & \qty{4.2}{\mega\byte} &  93\%  &&   \num{73554} & \num{14966} \\
    \textbf{Rotterdam}      && \qty{853} bldgs     & 2     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & 104\%  &&   \num{22246} &   \num{631} \\
    \textbf{Vienna}         && \qty{307} bldgs     & 2     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} &  89\%  &&   \num{47220} &  \num{2025} \\
    \textbf{Zürich}         && \qty{52834} bldgs   & 2     &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} &  89\%  && \num{3472989} &  \num{4069} \\
    \bottomrule
  \end{tabular}
    \begin{tablenotes}[flushleft]
      \footnotesize
      \item ${}^{\text{(a)}}$ `tex' is textures stored; `mat' is material stored
      \item ${}^{\text{(b)}}$ compression is $\frac{CityJSONSeq}{CityJSON}$, smaller is better
      \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
    \end{tablenotes}
  \end{threeparttable}
\end{table*}

With \emph{cjseq}, we have converted several publicly available files (details of files at \url{https://cityjson.org/datasets}), and Table~\ref{tab:stats} shows an overview of the files stored both in CityJSON and CityJSONSeq.
% TODO: setup website for hosting the files

First observe that---contrary to intuition---the size of a dataset, serialised to a JSON file, stored as a CityJSONSeq is around 12\% compacter than as a CityJSON, and in the case \textbf{3DBAG} it is 36\% compacter.
The main reason is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
For instance, the dataset \textbf{Helsinki} contains more than 3 millions vertices, but its largest feature contains only but 2202.
As a result, the CityJSONSeq stored in a file is only but 72\% of the original CityJSON file.
Only one dataset sees its size slightly increase when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
This is caused, we believe, by the fact that the dataset has textures that are shared by most of the buildings, and the \texttt{"appearance"} property must be repeated for each of the feature (in the CityJSON file it is only stored once and referenced by all buildings).

%%%
\subsection{Processing speed comparison}

We compare the speed and memory footprint of accessing each city object in a dataset.
This operation is common in applications that manipulate city models.

When a city model is stored in its entirety in a CityJSON object, we need to load the whole CityJSON object into memory in order to access the transform and vertices objects for instance.

With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one.
As shown in the experiments below, this allows a very efficient operation in terms of both CPU and memory usage.
Operations that would highly benefit from this are subsetting, merging a citymodel.
Things like EPSG reassignment or metadata updates wouldn't even require to loop through the features, just to alter the first object in the file.

We processed all the datasets from Table~\ref{tab:datasets} with 2 simple Python scripts that loops through each city objects and their geometries and increments a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script.} and the time used is shown in Table~\ref{tab:ramtime}.
\begin{table*}
  \centering
  \caption{}
  \small
  \begin{tabular}
    {@{}lcrrcrrr@{}}\toprule
    &&  \multicolumn{2}{c}{\textbf{RAM used (MB)}} && \multicolumn{3}{c}{\textbf{time (s)}} \\ 
    \cmidrule{3-4} \cmidrule{6-8} 
     && CityJSON & CityJSONSeq && CityJSON & CityJSONSeq & diff \\
    \midrule
     \textbf{3DBAG}         &&   76.9 & 16.1  &&   0.10 & 0.07 & 1.4 \\
     \textbf{Helsinki}      && 3743.1 & 15.0  &&  13.39 & 2.74 & 4.9 \\
     \textbf{Helsinki\_tex} && 5004.8 & 19.1  &&  29.60 & 4.72 & 6.3 \\
     \textbf{Ingolstadt}    &&   65.5 & 21.3  &&   0.08 & 0.06 & 1.3 \\
     \textbf{Montréal}      &&   79.3 & 20.8  &&   0.11 & 0.07 & 1.6 \\
     \textbf{NYC}           &&  949.5 & 16.0  &&   1.78 & 0.70 & 2.5 \\
     \textbf{Railway}       &&   69.6 & 29.6  &&   0.09 & 0.07 & 1.3 \\
     \textbf{Rotterdam}     &&   42.4 & 14.6  &&   0.04 & 0.04 & 1.0 \\
     \textbf{Vienna}        &&   60.1 & 15.7  &&   0.06 & 0.05 & 1.2 \\
     \textbf{Zurich}        && 2793.1 & 16.3  &&   6.05 & 2.00 & 3.0 \\
    \bottomrule
  \end{tabular}%
  \label{tab:ramtime}
\end{table*}

The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files.
At least for operations that do not require analysing or calculating city objects that are close to each other.
Operations like calculating volumes, merging and subsetting files, finding city objects having specific attributes, etc.\ will all use significantly less memory and be significantly faster.

%%%
%
\section{Discussion and future work}%
\label{sec:discussion}

\begin{itemize}
  \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
  \item it opens up many possibilities for processing large 3dcm by using Unix pipes. A few of the tools developed for CityJSON are already reading/writing to stdin/stdout, so they can be chained together.
  \item cjseqview, cjseqval are two of them, but developing others to eg calculate volumes or repair geometries would be trivial
  \item the nice part is that they tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
  \item extensions: these should be placed in the metadata or collection
\end{itemize}

Disadvantages:
\begin{itemize}
  \item     Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
  \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.
  \item  Appearances (esp. Material) are relatively heavy objects with many properties. If an appearance is shared across most CityObjects, duplicating it in each Feature could negate the effect of small vertex indices and actually increase the file size compared to a regular CityJSON file. This is just theorizing, haven't tested.
\end{itemize}

% TODO: add conclusions

{
	\begin{spacing}{1.17}
		\normalsize
		\bibliography{refs} % Include your own bibliography (*.bib), style is given in isprs.cls
	\end{spacing}
}

\end{document}

### 2024-04-18 scripts/load_cityjson.py

@@ -1,0 +1,26 @@
+# load_cityjson.py
+
+import json
+import sys
+import resource
+
+fpath = sys.argv[1]
+
+no_solid = 0
+no_msurface = 0
+with open(fpath, "r") as fo:
+    cm = json.load(fo)
+    for coid, co in cm["CityObjects"].items():
+        # process the CityObject with information from cm['transform']
+        # pass
+        for geom in co.get("geometry", []):
+            if geom["type"] == "Solid":
+                no_solid += 1
+            elif geom["type"] == "MultiSurface":
+                no_msurface += 1
+
+memory = resource.getrusage(resource.RUSAGE_SELF)
+print(memory.ru_maxrss / 1024 / 1024)
+
+# print(no_solid)
+# print(no_msurface)

### 2024-04-18 scripts/load_cityjsonseq.py

@@ -1,0 +1,38 @@
+# load_cityjsonseq.py
+
+import json
+from sys import argv
+import sys
+import psutil
+import resource
+
+fpath = argv[1]
+
+maxrss = 0.0
+no_solid = 0
+no_msurface = 0
+with open(fpath, "r") as fo:
+    meta = json.loads(fo.readline())
+    for feature_str in fo:
+        feature = json.loads(feature_str)
+        memory = resource.getrusage(resource.RUSAGE_SELF)
+        a = memory.ru_maxrss / 1024 / 1024
+        if a > maxrss:
+            maxrss = a
+        # process the feature with the information in 'meta', then discard
+        # pass
+        for coid, co in feature["CityObjects"].items():
+            for geom in co.get("geometry", []):
+                if geom["type"] == "Solid":
+                    no_solid += 1
+                elif geom["type"] == "MultiSurface":
+                    no_msurface += 1
+
+memory = resource.getrusage(resource.RUSAGE_SELF)
+a = memory.ru_maxrss / 1024 / 1024
+if a > maxrss:
+    maxrss = a
+
+print(maxrss)
+# print(no_solid)
+# print(no_msurcdface)

### 2024-04-18 scripts/size.py

@@ -1,0 +1,46 @@
+import sys
+import json
+import os
+import glob
+
+
+fs = glob.glob('../data/*.jsonl')
+
+# print(fs)
+# sys.exit()
+
+for each in fs:
+    # f = open('/Users/hugo/data/cityjson/3dcities/v2.0/3-20-DELFSHAVEN.city.jsonl')
+    f = open(each)
+    
+    totalkb = 0.0
+    maxchar = 0
+    maxv = 0
+    li = 0
+    for (i, line) in enumerate(f.readlines()):
+        kb = round(len(line) / 1024.0)
+        totalkb += kb
+        if kb > maxchar:
+            maxchar = kb
+            li = i
+        j = json.loads(line)
+        if len(j["vertices"]) > maxv:
+            maxv = len(j["vertices"])
+
+    # print("max:", maxchar, li)
+    # print("max:", maxchar)
+    # print("avg: {:.1f}".format(totalkb / (i+1.0)))
+    # f.seek(0)
+    # j2 = json.dumps(j, separators=(',', ':'))
+    # fo = open('out.jsonl', 'w')
+    # fo.write(j2)
+    # thesize = round(os.path.getsize("out.jsonl") / 1024.0)
+    # print(each, thesize)
+    print(each, maxchar, "{:.1f}".format(totalkb / (i+1.0)), maxv)
+
+fs = glob.glob('../data/*.json')
+for each in fs:
+    f = open(each)
+    j = json.loads(f.read())
+    print(each, len(j["vertices"]))
+

### 2024-04-18 scripts/test.sh

@@ -1,0 +1,39 @@
+printf "==3DBAG\n"
+\time python load_cityjson.py ../data/3DBAG.city.json
+\time python load_cityjsonseq.py ../data/3DBAG.city.jsonl
+
+printf "==Helsinki\n"
+\time python load_cityjson.py ../data/Helsinki.city.json
+\time python load_cityjsonseq.py ../data/Helsinki.city.jsonl
+
+printf "==Helsinki_tex\n"
+\time python load_cityjson.py ../data/Helsinki_tex.city.json
+\time python load_cityjsonseq.py ../data/Helsinki_tex.city.jsonl
+
+printf "==Ingolstadt\n"
+\time python load_cityjson.py ../data/Ingolstadt.city.json
+\time python load_cityjsonseq.py ../data/Ingolstadt.city.jsonl
+
+printf "==Montréal\n"
+\time python load_cityjson.py ../data/Montréal.city.json
+\time python load_cityjsonseq.py ../data/Montréal.city.jsonl
+
+printf "==NYC\n"
+\time python load_cityjson.py ../data/NYC.json
+\time python load_cityjsonseq.py ../data/NYC.jsonl
+
+printf "==Railway\n"
+\time python load_cityjson.py ../data/Railway.city.json
+\time python load_cityjsonseq.py ../data/Railway.city.jsonl
+
+printf "==Rotterdam\n"
+\time python load_cityjson.py ../data/Rotterdam.json
+\time python load_cityjsonseq.py ../data/Rotterdam.jsonl
+
+printf "==Vienna\n"
+\time python load_cityjson.py ../data/Vienna.city.json
+\time python load_cityjsonseq.py ../data/Vienna.city.jsonl
+
+printf "==Zurich\n"
+\time python load_cityjson.py ../data/Zurich.city.json
+\time python load_cityjsonseq.py ../data/Zurich.city.jsonl

### 2024-04-19 paper/main.tex
introductions

@@
 The main idea is to decompose a CityJSON file into its individual features (eg each Building, each Bridge, etc.) and create several JSON objects of a newly defined type \texttt{CityJSONFeature}.
-We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to/from CityJSONSeq, and discuss different aspects of the resulting CityJSONSeq, \eg\ size, usability, etc.
+We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ size, usability, etc.
 
-We think that CityJSONSeq is a better format than CityJSON
-CityJSONSeq is at about 10\% more compacter, takes an order of magnitude less time to process and uses significantly less memory (for some operations, not all).
+We consider CityJSONSeq to be a better format than CityJSON, since it is at about 10\% more compact, takes an order of magnitude less time to process and uses significantly less memory for certain operations.
 
@@
 
+%@HUGO several implementations or plugins?
+
 %
@@
 However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries (see Figure~\ref{fig:cj_idea}) made the \emph{streaming} of very large datasets complex, if not impossible.
-Because more and more datasets from cities are very large (it is not uncommon to have CityGML-GML files over 2GB), this made the processing of CityJSON difficult.
+Since city datasets are progressively becoming larger (it is not uncommon to have CityGML-GML files of over 2GB), the large data volume can hinder the processing of CityJSON files.
 As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.
@@
 
-We have implemented and released an open-source program to convert between in either direction between CityJSON and CityJSONSeq objects; we describe it in Section~\ref{sec:experiments}.
+We have implemented and released an open-source software to convert between CityJSON and CityJSONSeq objects; we describe it in Section~\ref{sec:experiments}.
 We also analyse the size of CityJSONSeq files versus CityJSON files, and that for several real-world datasets.
@@
 We discuss in Section~\ref{sec:experiments} the reasons for this interesting finding.
-Finally, we will also present the open-source software that are currently being developed to process streams of CityJSON datasets.
+Finally, we will also present the open-source software for processing streams of CityJSON datasets, that is currently being developed.
 

### 2024-04-19 paper/main.tex
structure

@@
   \item \texttt{"metadata"}: different metadata related to the dataset can be stored. The most important is the definition of the coordinate reference system (CRS)
-  \item \texttt{"transform"}: CityJSON \texttt{"vertices"} are compressed and stored with integers only. The parameters of this property allows us to convert from those to real-world coordinates.
-  \item \texttt{"CityObjects"}: a dictionary where the properties are the identifiers of the city objects (\emph{IDs}), which are any CityGML city objects (for instance a \texttt{Building}, a \texttt{BuildingPart}, a \texttt{SolitaryVegetationObject}, etc.).
-  Each of the city objects are listed one after the others, even if some are \texttt{"children"} of others 
+  \item \texttt{"transform"}: CityJSON \texttt{"vertices"} are compressed and stored as integers only.
+The parameters of this property allow us to convert from those integers back to real-world coordinates.
+  \item \texttt{"CityObjects"}: a dictionary where the properties are the identifiers of the city objects (\emph{IDs}), which are any CityGML city object (for instance a \texttt{Building}, a \texttt{BuildingPart}, a \texttt{SolitaryVegetationObject}, etc.).
+  Each of the city objects is listed one after the other, even if some are \texttt{"children"} of others.
   As an example, for a \texttt{Building} containing 2 parts, the 3 objects will be represented at the same level and linked by their \emph{IDs}, as shown in Figure~\ref{fig:parents_children}.
@@
   The fact that a dictionary is used means that developers have direct access to the city objects through their IDs (and also in constant time if a hashmap is used to implement the dictionary while parsing the file).
-  \item \texttt{"vertices"}: The 3D geometric primitives are those of CityGML, and multi/composite solids having several parts are supported.
-  A geometric primitive does not list all the coordinates of its vertices, rather the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and geometric primitives refer to the position of a vertex in that array.
-  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is reused, because it has been used for many years, with success, in the computer graphics community.
+  \item \texttt{"vertices"}: The 3D geometric primitives are those of CityGML, and multi/composite solids with several parts are supported.
+  A geometric primitive does not list all the coordinates of its vertices, rather the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and the geometric primitives refer to the position of a vertex in that array.
+  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is reused, because it has been successfully used for many years in the computer graphics community.
   There are several advantages to this approach.
-  First, the files can be compressed: 3D vertices are often shared by several surfaces, and repeating them can be costly (especially if they are very precise, often sub-millimetre is used).
-  Second, this increases the topological relationships that are explicitly stored in the file, and several operations (\eg\ are 2 buildings adjacent?) can be sped up and made more robust.
+  First, the files can be compressed: 3D vertices are often shared by several surfaces, and repeating them can be costly, especially if they are very precise (sub-millimetre precision is often used).
+  Second, this approach increases the topological relationships that are explicitly stored in the file, and several operations (\eg\ are 2 buildings adjacent?) can be sped up and made more robust.
   Third, it is very easy to convert to a representation listing all coordinates; the inverse is not true. 
-  However, this is where the streaming of geometries is problematic: the array of "vertices" can for very large areas, contains several millions vertices.
-  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can be for streaming first waiting for millions of unused vertices to be transferred.
+  However, this is where the streaming of geometries is problematic: the array of "vertices" contains several millions vertices, especially for very large areas, .
+  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which means waiting for millions of unused vertices to be loaded.
   \item \texttt{"appearances"}: Both textures and materials for surfaces are supported.  
@@
 \end{lstlisting}
-  \caption{CityJSON mechanism to flatten out the schema: each city object are stored in a flat list, and they are linked together with the properties \texttt{"parents"} and \texttt{"children"}.}%
+  \caption{CityJSON mechanism to flatten out the schema: the city objects are stored in a flat list, and they are linked together with the properties \texttt{"parents"} and \texttt{"children"}.}%
 \label{fig:parents_children}
@@
 The presence of \emph{XLinks} means that references between elements are possible (\eg\ reusing surfaces), but if in the stream the referenced surface has not appeared, the receiver cannot process the information.
-It is not always possible to order a CityGML files so that all the references can be resolved without having to store locally information until it appears in the stream.
+It is not always possible to order CityGML files so that all the references can be resolved without having to store locally information until it appears in the stream.
 It is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.
+%@HUGO: I don't understand the last 2 sentenses.
 
@@
 
-For formats that use a global indexing of vertices (as written above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
+For formats that use a global indexing of vertices (as mentioned above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
 \citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that vertices are not to be used anymore (and thus can be freed from memory).

### 2024-04-19 paper/main.tex
:

@@
 CityJSONSeq was added to the CityJSON version 2.0 standard that was recently released, and it allows us to stream very large 3D city models.
-The main idea is to decompose a CityJSON file into its individual features (eg each Building, each Bridge, etc.) and create several JSON objects of a newly defined type \texttt{CityJSONFeature}.
+The main idea is to decompose a CityJSON file into its individual features (eg each Building, each Bridge, etc.) and create several JSON objects of a newly defined type: \texttt{CityJSONFeature}.
 We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ size, usability, etc.

### 2024-04-19 paper/main.tex

@@
 The rule \#4 ensures that those are available. 
-The CityJSON object must contain \texttt{"transform"} and eventually the other properties if needed; \texttt{"CityObjects"} and \texttt{"vertices"} must be present but be empty (to ensure that the JSON object is valid).
+The CityJSON object must contain \texttt{"transform"} and eventually the other properties if needed; \texttt{"CityObjects"} and \texttt{"vertices"} must be present but they must be empty (to ensure that the JSON object is valid).
 
@@
 
-First observe that---contrary to intuition---the size of a dataset, serialised to a JSON file, stored as a CityJSONSeq is around 12\% compacter than as a CityJSON, and in the case \textbf{3DBAG} it is 36\% compacter.
-The main reason is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
+First observe that---contrary to intuition---the size of a dataset, serialised to a JSON file, stored as a CityJSONSeq is around 12\% more compact than as a CityJSON, and in the case of \textbf{3DBAG} it is 36\% compacter.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
+%@Hugo: Sentence above not super clear. 
 For instance, the dataset \textbf{Helsinki} contains more than 3 millions vertices, but its largest feature contains only but 2202.
 As a result, the CityJSONSeq stored in a file is only but 72\% of the original CityJSON file.
-Only one dataset sees its size slightly increase when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
+Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
 This is caused, we believe, by the fact that the dataset has textures that are shared by most of the buildings, and the \texttt{"appearance"} property must be repeated for each of the feature (in the CityJSON file it is only stored once and referenced by all buildings).
@@
 With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one.
-As shown in the experiments below, this allows a very efficient operation in terms of both CPU and memory usage.
+As shown in the experiments below, this allows for a very efficient operation in terms of both CPU and memory usage.
 Operations that would highly benefit from this are subsetting, merging a citymodel.
@@
 
-We processed all the datasets from Table~\ref{tab:datasets} with 2 simple Python scripts that loops through each city objects and their geometries and increments a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
-The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script.} and the time used is shown in Table~\ref{tab:ramtime}.
+We processed all the datasets from Table~\ref{tab:datasets} with 2 simple Python scripts that loop through the city objects and their geometries and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
+The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script.} and the time used are shown in Table~\ref{tab:ramtime}.
 \begin{table*}
@@
 
-The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files.
-At least for operations that do not require analysing or calculating city objects that are close to each other.
-Operations like calculating volumes, merging and subsetting files, finding city objects having specific attributes, etc.\ will all use significantly less memory and be significantly faster.
+The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files, at least for operations that do not require analysing or calculating city objects that are close to each other.
+Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.
 

### 2024-04-19 paper/main.tex
Try to improve unclear sentences

@@
   \item it was adopted by the Dutch government as a 3D standard to distribute nation-wide 3D datasets; 
-  \item several implementations, \eg\ FME, were developed.
+  \item several implementations and plugins have been developed, most notably FME\@.
 \end{enumerate}
-
-%@HUGO several implementations or plugins?
 
@@
 However, notice that this is only true for simple files and GML features stored following the Simple Feature paradigm~\citep{OGC-SF}.
-For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used (for reusing geometries and for storing some topological relationships), the conversion often needs a large amount of pre-processing.
-The presence of \emph{XLinks} means that references between elements are possible (\eg\ reusing surfaces), but if in the stream the referenced surface has not appeared, the receiver cannot process the information.
-It is not always possible to order CityGML files so that all the references can be resolved without having to store locally information until it appears in the stream.
-It is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.
-%@HUGO: I don't understand the last 2 sentenses.
+For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used, the conversion often needs a large amount of pre-processing.
+
+\emph{XLinks} are links links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building for instance).
+If in the stream the linked surface has not appeared yet, the receiver cannot process the cube and needs to wait for that specific surface to appear.
+It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store locally information until it appears in the stream.
+However, it is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.
 
@@
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
-%@Hugo: Sentence above not super clear. 
 For instance, the dataset \textbf{Helsinki} contains more than 3 millions vertices, but its largest feature contains only but 2202.
+The fact that many indices are used for geometries (and for textures) means that if large numbers are used then the size of the file will grow.
 As a result, the CityJSONSeq stored in a file is only but 72\% of the original CityJSON file.
@@
 \begin{itemize}
-  \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
+  \item CityJSONSeq is an attractive alternative to CityJSON; in most uses it is more compact, its processing requires less memory and it is faster.
   \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
@@
 \begin{itemize}
-  \item     Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
+  \item  Software implementing CityJSON needs to account for both regular CityJSON and CityJSONSeq.
   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.

### 2024-04-19 paper/main.tex
Try to improve unclear sentences

@@
 \begin{itemize}
-  \item CityJSONSeq is an attractive alternative to CityJSON; in most uses it is more compact, its processing requires less memory and it is faster.
+  \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
   \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
@@
 \begin{itemize}
-  \item  Software implementing CityJSON needs to account for both regular CityJSON and CityJSONSeq.
+  \item     Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.

### 2024-04-19 paper/main.tex
Fix text about xlinks

@@
 For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used, the conversion often needs a large amount of pre-processing.
-
-\emph{XLinks} are links links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building for instance).
-If in the stream the linked surface has not appeared yet, the receiver cannot process the cube and needs to wait for that specific surface to appear.
+\emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building, for instance).
+If in the stream the linked surface has not appeared yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear.
 It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store locally information until it appears in the stream.
-However, it is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.
+However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is copy the geometry of the 6th surface to the cube), but this means that its size is increased, and that converting back to the original file will not be possible.
 

### 2024-04-19 paper/main.tex
discussion

@@
   \item it was adopted by the Dutch government as a 3D standard to distribute nation-wide 3D datasets; 
-  \item several implementations and plugins have been developed, most notably FME\@.
+  \item several implementations, \eg\ FME, were developed.
 \end{enumerate}
+
+%@HUGO several implementations or plugins?
 
@@
 However, notice that this is only true for simple files and GML features stored following the Simple Feature paradigm~\citep{OGC-SF}.
-For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used, the conversion often needs a large amount of pre-processing.
-\emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building, for instance).
-If in the stream the linked surface has not appeared yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear.
-It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store locally information until it appears in the stream.
-However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is copy the geometry of the 6th surface to the cube), but this means that its size is increased, and that converting back to the original file will not be possible.
+For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used (for reusing geometries and for storing some topological relationships), the conversion often needs a large amount of pre-processing.
+The presence of \emph{XLinks} means that references between elements are possible (\eg\ reusing surfaces), but if in the stream the referenced surface has not appeared, the receiver cannot process the information.
+It is not always possible to order CityGML files so that all the references can be resolved without having to store locally information until it appears in the stream.
+It is always possible to resolve the \emph{XLinks} before streaming a dataset, but this means that its size is increased, and that converting back to the original file will not be possible.
+%@HUGO: I don't understand the last 2 sentenses.
 
@@
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
+%@Hugo: Sentence above not super clear. 
 For instance, the dataset \textbf{Helsinki} contains more than 3 millions vertices, but its largest feature contains only but 2202.
-The fact that many indices are used for geometries (and for textures) means that if large numbers are used then the size of the file will grow.
 As a result, the CityJSONSeq stored in a file is only but 72\% of the original CityJSON file.

### 2024-04-19 paper/main.tex
A few fixes + adding GeoJSON in RW

--- 2024-04-19 paper/main.tex
@@
 CityJSONSeq was added to the CityJSON version 2.0 standard that was recently released, and it allows us to stream very large 3D city models.
-The main idea is to decompose a CityJSON file into its individual features (eg each Building, each Bridge, etc.) and create several JSON objects of a newly defined type: \texttt{CityJSONFeature}.
+The main idea is to decompose a CityJSON dataset into its individual features (eg each Building, each Bridge, etc.) and create several indepedent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
 We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ size, usability, etc.
-
-We consider CityJSONSeq to be a better format than CityJSON, since it is at about 10\% more compact, takes an order of magnitude less time to process and uses significantly less memory for certain operations.
-
+We consider CityJSONSeq to be a better format than CityJSON because: (1) once serialised it is about 10\% more compact; (2) it takes an order of magnitude less time to process; and (3) it uses significantly less memory for certain operations.
 }
@@
 
+For the \emph{GeoJSON} format~\citep{IETF-GeoJSON}, which follows the Simple Feature paradigm, creating a stream is trivial since each of the features in the dataset becomes one JSON object and one line~\citep{IETF-GeoJSONSeq}.
+There are no links possible between features, they are all independent.
+
+%
+
 For formats that use a global indexing of vertices (as mentioned above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
@@
 
-With \emph{cjseq}, we have converted several publicly available files (details of files at \url{https://cityjson.org/datasets}), and Table~\ref{tab:stats} shows an overview of the files stored both in CityJSON and CityJSONSeq.
+With \emph{cjseq}, we have converted several publicly available files (details of files at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}), and Table~\ref{tab:stats} shows an overview of the files stored both in CityJSON and CityJSONSeq.
 % TODO: setup website for hosting the files
 
-First observe that---contrary to intuition---the size of a dataset, serialised to a JSON file, stored as a CityJSONSeq is around 12\% more compact than as a CityJSON, and in the case of \textbf{3DBAG} it is 36\% compacter.
-The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and no not increase to several millions.
-For instance, the dataset \textbf{Helsinki} contains more than 3 millions vertices, but its largest feature contains only but 2202.
-The fact that many indices are used for geometries (and for textures) means that if large numbers are used then the size of the file will grow.
-As a result, the CityJSONSeq stored in a file is only but 72\% of the original CityJSON file.
+First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{3DBAG} it is 36\% compacter.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to several millions.
+For instance, the dataset \textbf{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
+The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the size of the file will grow; if they maximum vertex index is around 2000 for each feature, then the file size will be less than the original file.
 Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
@@
 
-When a city model is stored in its entirety in a CityJSON object, we need to load the whole CityJSON object into memory in order to access the transform and vertices objects for instance.
+When a city model is stored in its entirety in one CityJSON object, we need to load the whole CityJSON object into memory in order to access the \texttt{"transform"} and \texttt{"vertices"} properties for instance.
 
 With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one.
-As shown in the experiments below, this allows for a very efficient operation in terms of both CPU and memory usage.
-Operations that would highly benefit from this are subsetting, merging a citymodel.
-Things like EPSG reassignment or metadata updates wouldn't even require to loop through the features, just to alter the first object in the file.
+As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
+Operations that would highly benefit from this are subsetting and merging a citymodel.
+Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
 
-We processed all the datasets from Table~\ref{tab:datasets} with 2 simple Python scripts that loop through the city objects and their geometries and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
+%
+
+We have processed all the datasets from Table~\ref{tab:datasets} with two simple Python scripts that loop through the city objects and their geometries, and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
+This is just an example of a simple local operation, any other operation such as calculating the area of the façades or counting the number of windows could have been performed.
 The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script.} and the time used are shown in Table~\ref{tab:ramtime}.
+The scripts are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/scripts/}.
 \begin{table*}
   \centering
-  \caption{}
+  \caption{Comparison of the processing time and max RAM use for processing the same data stored in a CityJSON file and a CityJSONSeq file.}
   \small
@@
 \begin{itemize}
+  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
-  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item it opens up many possibilities for processing large 3dcm by using Unix pipes.
 A few of the tools developed for CityJSON are already reading/writing to stdin/stdout, so they can be chained together.
-  \item cjseqview, cjseqval are two of them, but developing others to eg calculate volumes or repair geometries would be trivial
-  \item the nice part is that they tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
+  \item cjseqview, cjseqval are two of them, but developing others to eg filter, calculate volumes or repair geometries would be trivial
+  \item the nice part is that the tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
   \item extensions: these should be placed in the metadata or collection

### 2024-04-19 paper/main.tex
Add why rotterdam size increased, I tested with a dataset

@@
 The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the size of the file will grow; if they maximum vertex index is around 2000 for each feature, then the file size will be less than the original file.
+
+%
+
 Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
-This is caused, we believe, by the fact that the dataset has textures that are shared by most of the buildings, and the \texttt{"appearance"} property must be repeated for each of the feature (in the CityJSON file it is only stored once and referenced by all buildings).
+This is caused, primarily, by the fact that the dataset has textures that are shared by most of the buildings, and the \texttt{"appearance"} property must be repeated for each of the feature (in the CityJSON file it is only stored once and referenced by all buildings).
+If we remove the textures from the CityJSON file it is \qty{2.1}{\mega\byte} and its CityJSONSeq equivalent is \qty{1.3}{\mega\byte} (which means a difference of 62\%)
 
@@
   \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
+  \item CityJSONSeq is an attractive alternative to CityJSON; in most uses it is more compact, its processing requires less memory and it is faster.
+  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item it opens up many possibilities for processing large 3dcm by using Unix pipes.

### 2024-04-22 paper/main.tex
Add why rotterdam size increased, I tested with a dataset
We need to think about the title I guess
merge

@@
 
+% TODO: the title is perhaps too short? Or is good since it's punchy?
 \title{Streaming CityJSON datasets}
+% \title{CityJSONSeq: streaming massive CityJSON datasets}
+% \title{Streaming massive CityJSON datasets with CityJSONSeq}
 
@@
 \begin{itemize}
-  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
-  \item CityJSONSeq is an attractive alternative to CityJSON; in most uses it is more compact, its processing requires less memory and it is faster.
   \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
@@
 \begin{itemize}
-  \item  Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
+  \item     Software implementing CityJSON needs to account for both regular CityJSON and CityJSONSeq.
   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.

### 2024-04-22 scripts/shared_vertices.py
Count shared vertices

@@ -1,0 +1,39 @@
+"""
+Script for counting the shared vertices.
+It writes a CSV file with the shared vertex count for each vertex, to the same
+directory as the input file.
+"""
+import json
+import sys
+from pathlib import Path
+import sys
+import csv
+
+if __name__ == '__main__':
+    inpath = Path(sys.argv[1]).resolve()
+
+    with inpath.open("r") as f:
+        data = json.load(f)
+
+    vertex_counts = {i: 0 for i in range(len(data['vertices']))}
+
+    for co_id, co in data["CityObjects"].items():
+        for geometry in co["geometry"]:
+            for surface in geometry["boundaries"]:
+                for ring in surface:
+                    for vertex in ring:
+                        vertex_counts[vertex] += 1
+
+    vertex_counts_sorted = dict(sorted(vertex_counts.items(), key=lambda item: item[1], reverse=True))
+
+    nr_shared = 0
+    with inpath.parent.joinpath(inpath.stem).with_suffix(".csv").open("w") as f:
+        csv_writer = csv.writer(f)
+        csv_writer.writerow(["Vertex", "Count"])
+        for vertex, count in vertex_counts_sorted.items():
+            if count > 1:
+                nr_shared += 1
+                csv_writer.writerow([vertex, count])
+
+    print(f"Nr. vertices: {len(vertex_counts)}")
+    print(f"Nr. shared vertices: {nr_shared}")

### 2024-04-22 scripts/shared_vertices.py
Count shared vertices between cityobjects

@@ -1,5 +1,5 @@
 """
-Script for counting the shared vertices.
+Script for counting how many CityObjects share a shared vertex.
 It writes a CSV file with the shared vertex count for each vertex, to the same
 directory as the input file.
 """
@@ -15,25 +15,33 @@ if __name__ == '__main__':
     with inpath.open("r") as f:
         data = json.load(f)
 
-    vertex_counts = {i: 0 for i in range(len(data['vertices']))}
+    vertex_counts = {i: set() for i in range(len(data['vertices']))}
 
     for co_id, co in data["CityObjects"].items():
         for geometry in co["geometry"]:
             for surface in geometry["boundaries"]:
                 for ring in surface:
                     for vertex in ring:
-                        vertex_counts[vertex] += 1
+                        vertex_counts[vertex].add(co_id)
 
-    vertex_counts_sorted = dict(sorted(vertex_counts.items(), key=lambda item: item[1], reverse=True))
+    vertex_counts_sorted = dict(sorted(vertex_counts.items(), key=lambda item: len(item[1]), reverse=True))
 
     nr_shared = 0
+    size_vertex = sys.getsizeof(data['vertices'][0])
+    total_size_vertices = size_vertex * len(data['vertices'])
+    total_size_vertices_unshared = 0
     with inpath.parent.joinpath(inpath.stem).with_suffix(".csv").open("w") as f:
         csv_writer = csv.writer(f)
         csv_writer.writerow(["Vertex", "Count"])
-        for vertex, count in vertex_counts_sorted.items():
-            if count > 1:
+        for vertex, ids in vertex_counts_sorted.items():
+            if len(ids) > 1:
                 nr_shared += 1
-                csv_writer.writerow([vertex, count])
+                total_size_vertices_unshared += size_vertex * len(ids)
+                csv_writer.writerow([vertex, len(ids)])
+            else:
+                total_size_vertices_unshared += size_vertex
 
     print(f"Nr. vertices: {len(vertex_counts)}")
     print(f"Nr. shared vertices: {nr_shared}")
+    print(f"Memory size vertices: {total_size_vertices} bytes")
+    print(f"Memory size vertices if none of them are shared: {total_size_vertices_unshared} bytes")

### 2024-04-22 scripts/shared_vertices_2.py
Add cheap code to check how many shared vertices

@@ -1,0 +1,29 @@
+
+import json
+import sys
+from pathlib import Path
+import sys
+import csv
+
+
+inpath1 = Path("/Users/hugo/data/cityjson/3dcities/v2.0/VM05_2009.city.json").resolve()
+# inpath1 = Path("/Users/hugo/temp/r/Rotterdam_wo_tex.json").resolve()
+with inpath1.open("r") as f:
+    j = json.load(f)
+v_json = len(j["vertices"])
+
+inpath2 = Path("/Users/hugo/data/cityjson/3dcities/v2.0/VM05_2009.city.jsonl").resolve()
+# inpath2 = Path("/Users/hugo/temp/r/Rotterdam_wo_tex.jsonl").resolve()
+f = open(inpath2)
+v_jsonl = 0
+for (i, line) in enumerate(f.readlines()):
+    j2 = json.loads(line)
+    v_jsonl += len(j2["vertices"]) 
+            
+
+# print(v_json)
+print(v_jsonl / v_json)
+print(v_json / v_jsonl)
+
+
+

### 2024-04-23 paper/main.tex
Add conversion time as indication

@@
 
+It should be observe that the conversion is an efficient process: the rather large dataset (\textbf{Helskinki}, which contains more than \qty{77000} buildings and whose CityJSON file is \qty{572}{\mega\byte}) takes only \qty{4.7}{\sec} to be converted to a CityJSONSeq file, and the reverse operation takes \qty{5.7}{\sec} on a standard laptop.
+
 %%%
@@
 \begin{itemize}
+  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
-  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
   \item it opens up many possibilities for processing large 3dcm by using Unix pipes.
@@
 \begin{itemize}
-  \item     Software implementing CityJSON needs to account for both regular CityJSON and CityJSONSeq.
+  \item  Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.

### 2024-04-24 scripts/shared_vertices_2.py
Fix script to calculate shared vertices

@@ -4,26 +4,25 @@ import sys
 from pathlib import Path
 import sys
 import csv
+import glob
 
+fs = glob.glob('/Users/hugo/data/cityjson/CityJSONSeq-demo-files/*.json')
 
-inpath1 = Path("/Users/hugo/data/cityjson/3dcities/v2.0/VM05_2009.city.json").resolve()
-# inpath1 = Path("/Users/hugo/temp/r/Rotterdam_wo_tex.json").resolve()
-with inpath1.open("r") as f:
-    j = json.load(f)
-v_json = len(j["vertices"])
+for each in fs:
+    #-- json
+    with Path(each).resolve().open("r") as f:
+        j = json.load(f)
+    v_json = len(j["vertices"])
 
-inpath2 = Path("/Users/hugo/data/cityjson/3dcities/v2.0/VM05_2009.city.jsonl").resolve()
-# inpath2 = Path("/Users/hugo/temp/r/Rotterdam_wo_tex.jsonl").resolve()
-f = open(inpath2)
-v_jsonl = 0
-for (i, line) in enumerate(f.readlines()):
-    j2 = json.loads(line)
-    v_jsonl += len(j2["vertices"]) 
-            
+    #-- jsonl
+    eachl = each + "l"
+    f = open(eachl)
+    v_jsonl = 0
+    for (i, line) in enumerate(f.readlines()):
+        j2 = json.loads(line)
+        v_jsonl += len(j2["vertices"]) 
 
-# print(v_json)
-print(v_jsonl / v_json)
-print(v_json / v_jsonl)
+    print(each, v_jsonl / v_json)
 
 
 

### 2024-04-24 paper/main.tex
Update size of datasets + shared vertices

@@
   \begin{tabular}
-    {@{}lcccccrrrcrr@{}}\toprule
-    &&  \multicolumn{3}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{2}{c}{\textbf{vertices}}   \\ 
-    \cmidrule{3-5} \cmidrule{7-9} \cmidrule{11-12} 
-     && CityObjects &  LoDs & appearance\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ \\
+    {@{}lcccccrrrcrrr@{}}\toprule
+    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{3}{c}{\textbf{vertices}}   \\ 
+    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-12} 
+     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared \\
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs & 0+1.2+1.3+2.2 & &&    \qty{6.7}{\mega\byte} & \qty{4.3}{\mega\byte} &  64\%  &&   \num{82509} &  \num{4112} \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   & 1+2   &         && \qty{600}{\mega\byte} & \qty{432}{\mega\byte} &  72\%  && \num{3038576} &  \num{2202} \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & 1+2   & tex     && \qty{748}{\mega\byte} & \qty{675}{\mega\byte} &  90\%  && \num{3038576} &  \num{2202} \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      & 3     &         && \qty{5.1}{\mega\byte} & \qty{4.0}{\mega\byte} &  78\%  &&   \num{87972} & \num{12800} \\
-    \textbf{Montréal}       && \qty{294} bldgs     & 2     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} &  85\%  &&   \num{31585} &  \num{3393} \\
-    \textbf{NYC}            && \qty{23777} bldgs   & 2     &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} &  90\%  && \num{1035804} &  \num{2608} \\
-    \textbf{Railway}        && \qty{50} diff       & 3     & tex+mat && \qty{4.5}{\mega\byte} & \qty{4.2}{\mega\byte} &  93\%  &&   \num{73554} & \num{14966} \\
-    \textbf{Rotterdam}      && \qty{853} bldgs     & 2     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & 104\%  &&   \num{22246} &   \num{631} \\
-    \textbf{Vienna}         && \qty{307} bldgs     & 2     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} &  89\%  &&   \num{47220} &  \num{2025} \\
-    \textbf{Zürich}         && \qty{52834} bldgs   & 2     &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} &  89\%  && \num{3472989} &  \num{4069} \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} &  88\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
+    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} &  84\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} &  72\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} &  90\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} &  79\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} &  85\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} &  90\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
+    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  93\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
+    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & 104\%  &&     \num{22246} &     \num{631} & 20.0\% \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} &  89\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} &  89\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
     \bottomrule
@@
       \footnotesize
-      \item ${}^{\text{(a)}}$ `tex' is textures stored; `mat' is material stored
+      \item ${}^{\text{(a)}}$ appearance: `tex' is textures stored; `mat' is material stored
       \item ${}^{\text{(b)}}$ compression is $\frac{CityJSONSeq}{CityJSON}$, smaller is better

### 2024-04-24 paper/main.tex
Add 3DBV to the dataset

@@
      \textbf{3DBAG}         &&   76.9 &  16.1  &&   0.10 & 0.07 & 1.4 \\
+     \textbf{3DBV}          && 4101.8 & 123.8  &&  10.95 & 3.59 & 3.1 \\
      \textbf{Helsinki}      && 3743.1 &  15.0  &&  13.39 & 2.74 & 4.9 \\
@@
 
-The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files, at least for operations that do not require analysing or calculating city objects that are close to each other.
+The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files, at least for operations that do not require analysing or processing city objects that are close to each other.
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.

### 2024-04-24 scripts/test.sh
forgot to commit this

@@ -2,6 +2,10 @@ printf "==3DBAG\n"
 \time python load_cityjson.py ../data/3DBAG.city.json
 \time python load_cityjsonseq.py ../data/3DBAG.city.jsonl
 
+printf "==3DBV\n"
+\time python load_cityjson.py ../data/3DBV.city.json
+\time python load_cityjsonseq.py ../data/3DBV.city.jsonl
+
 printf "==Helsinki\n"
 \time python load_cityjson.py ../data/Helsinki.city.json
 \time python load_cityjsonseq.py ../data/Helsinki.city.jsonl

### 2024-04-24 scripts/wipeattribute.py
Add script to remove attributes from a file

@@ -1,0 +1,16 @@
+# load_cityjson.py
+
+import json
+import sys
+import resource
+
+fpath = sys.argv[1]
+
+with open(fpath, "r") as fo:
+    cm = json.load(fo)
+    for coid, co in cm["CityObjects"].items():
+        del cm["CityObjects"][coid]["attributes"]
+
+json_str = json.dumps(cm, separators=(',',':'))
+with open("out.json", "w") as foo:
+    foo.write(json_str)

### 2024-04-25 scripts/generate.py
Add script to generate CityJSON file with random cubes

@@ -1,0 +1,50 @@
+import json
+import sys
+import random
+
+
+def gen_onecube(offset):
+    g = {"type": "Solid", "lod": "1.1", "boundaries": []}
+    i = offset
+    b = []
+    b.append([ [i+0, i+3, i+2, i+1] ])
+    b.append([ [i+4, i+5, i+6, i+7] ])
+    b.append([ [i+0, i+1, i+5, i+4] ])
+    b.append([ [i+1, i+2, i+6, i+5] ])
+    b.append([ [i+2, i+3, i+7, i+6] ])
+    b.append([ [i+0, i+4, i+7, i+3] ])
+    g["boundaries"].append(b)
+
+    return g
+
+def gen_vertices(total):
+    vs = []
+    for i in range(total):
+        vs.append([random.randint(0, 10000), random.randint(0, 10000), random.randint(0, 10000)])
+    return vs
+
+
+def main():
+    number_of_co = 5000000
+    j = {"type": "CityJSON", "version": "2.0", "CityObjects": {}, "transform": {}}
+    j["transform"]["scale"] = [0.01, 0.01, 0.01]
+    j["transform"]["translate"] = [85000.1, 445006.2, 0.1]
+    offset = 0
+    for i in range(number_of_co):
+        cgeom = gen_onecube(offset)
+        offset += 8
+        coj = {"type": "Building", "geometry": []}
+        coj["geometry"].append(cgeom)
+        coid = "id_" + str(i)
+        j["CityObjects"][coid] = coj
+    j["vertices"] = gen_vertices(number_of_co * 8) 
+
+    json_str = json.dumps(j, separators=(',',':'))
+    # json_str = json.dumps(j, indent=2)
+    fname = "gen_" + str(int(number_of_co / 1000)) + "k.json"
+    with open(fname, "w") as foo:
+        foo.write(json_str)
+
+
+if __name__ == '__main__':
+    main()

### 2024-04-25 paper/main.tex
Change compression to a % from 0, simpler + clearer

@@
     \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-12} 
-     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared \\
+     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared\footnotesize ${}^{\text{(d)}}$ \\
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} &  88\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
-    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} &  84\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} &  72\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} &  90\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} &  79\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
-    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} &  85\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
-    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} &  90\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
-    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  93\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
-    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & 104\%  &&     \num{22246} &     \num{631} & 20.0\% \\
-    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} &  89\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
-    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} &  89\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 14\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
+    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 19\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 39\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 11\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 26\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 17\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 11\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
+    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
+    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 13\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 13\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
     \bottomrule
@@
       \item ${}^{\text{(a)}}$ appearance: `tex' is textures stored; `mat' is material stored
-      \item ${}^{\text{(b)}}$ compression is $\frac{CityJSONSeq}{CityJSON}$, smaller is better
+      \item ${}^{\text{(b)}}$ compression factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$
       \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
+      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects (\eg\ the vertices of a wall shared by two buildings)
     \end{tablenotes}
@@
 With \emph{cjseq}, we have converted several publicly available files (details of files at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}), and Table~\ref{tab:stats} shows an overview of the files stored both in CityJSON and CityJSONSeq.
+
 % TODO: setup website for hosting the files
@@
     \midrule
-     \textbf{3DBAG}         &&   76.9 &  16.1  &&   0.10 & 0.07 & 1.4 \\
-     \textbf{3DBV}          && 4101.8 & 123.8  &&  10.95 & 3.59 & 3.1 \\
-     \textbf{Helsinki}      && 3743.1 &  15.0  &&  13.39 & 2.74 & 4.9 \\
-     \textbf{Helsinki\_tex} && 5004.8 &  19.1  &&  29.60 & 4.72 & 6.3 \\
-     \textbf{Ingolstadt}    &&   65.5 &  21.3  &&   0.08 & 0.06 & 1.3 \\
-     \textbf{Montréal}      &&   79.3 &  20.8  &&   0.11 & 0.07 & 1.6 \\
-     \textbf{NYC}           &&  949.5 &  16.0  &&   1.78 & 0.70 & 2.5 \\
-     \textbf{Railway}       &&   69.6 &  29.6  &&   0.09 & 0.07 & 1.3 \\
-     \textbf{Rotterdam}     &&   42.4 &  14.6  &&   0.04 & 0.04 & 1.0 \\
-     \textbf{Vienna}        &&   60.1 &  15.7  &&   0.06 & 0.05 & 1.2 \\
-     \textbf{Zurich}        && 2793.1 &  16.3  &&   6.05 & 2.00 & 3.0 \\
+     \textbf{3DBAG}         &&   76.9 &  16.1  &&   0.10 & 0.07 & 1.4X \\
+     \textbf{3DBV}          && 4101.8 & 123.8  &&  10.95 & 3.59 & 3.1X \\
+     \textbf{Helsinki}      && 3743.1 &  15.0  &&  13.39 & 2.74 & 4.9X \\
+     \textbf{Helsinki\_tex} && 5004.8 &  19.1  &&  29.60 & 4.72 & 6.3X \\
+     \textbf{Ingolstadt}    &&   65.5 &  21.3  &&   0.08 & 0.06 & 1.3X \\
+     \textbf{Montréal}      &&   79.3 &  20.8  &&   0.11 & 0.07 & 1.6X \\
+     \textbf{NYC}           &&  949.5 &  16.0  &&   1.78 & 0.70 & 2.5X \\
+     \textbf{Railway}       &&   69.6 &  29.6  &&   0.09 & 0.07 & 1.3X \\
+     \textbf{Rotterdam}     &&   42.4 &  14.6  &&   0.04 & 0.04 & 1.0X \\
+     \textbf{Vienna}        &&   60.1 &  15.7  &&   0.06 & 0.05 & 1.2X \\
+     \textbf{Zurich}        && 2793.1 &  16.3  &&   6.05 & 2.00 & 3.0X \\
     \bottomrule

### 2024-04-25 paper/main.tex
WIP for explaining the compression factor

@@
 
-First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{3DBAG} it is 36\% compacter.
+First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{Helsinki} it is 39\% compacter.
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to several millions.
@@
 
-Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file (\textbf{Rotterdam}).
-This is caused, primarily, by the fact that the dataset has textures that are shared by most of the buildings, and the \texttt{"appearance"} property must be repeated for each of the feature (in the CityJSON file it is only stored once and referenced by all buildings).
-If we remove the textures from the CityJSON file it is \qty{2.1}{\mega\byte} and its CityJSONSeq equivalent is \qty{1.3}{\mega\byte} (which means a difference of 62\%)
+Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file: \textbf{Rotterdam}.
+Its CityJSONSeq size is 4\% larger than that of its CityJSON file.
+One reason for this is that the size of the CityJSON file is rather small in the first place, and therefore the indices are not high values.
+Having several features means that many JSON properties must be repeated for each feature, and this can cause the file size to slightly increase.
+
+%
+
+As an experiment, we have created around 125 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated; Figure~\ref{fig:synthetic_compression} shows the results.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
+  \caption{Compression factor of CityJSONSeq for different synthetic datasets containing only but simple cuboid buildings.}%
+\label{fig:synthetic_compression}
+\end{figure}
+Notice that as the filesize grows, the compression factor ($\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$) grows.
+The smallest file contains only but 526 buildings and its compression factor is -2\%, while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
+
+%
+
+% We can expect
 

### 2024-04-25 scripts/convert.py

@@ -1,0 +1,56 @@
+import sys
+import json
+import os
+import glob
+import matplotlib.pyplot as plt
+import matplotlib.ticker as ticker
+import matplotlib
+import numpy as np
+
+
+
+#-- 1. convert CityJSON => CityJSONSeq
+# fs = glob.glob('*.json')
+# for each in fs:
+#     # cjseq cat -f out.json > out.jsonl
+#     cmd = "cjseq cat -f {} > {}l".format(each, each)
+#     print(cmd)
+#     os.system(cmd)
+
+
+#-- 2. sizes
+fs = glob.glob('./data/*.json')
+nocubes = []
+scj = []
+scjseq = []
+for f in fs:
+    # print(f)
+    nocubes.append(int(f[f.find("_")+1:f.rfind(".")]))
+    scj.append(os.path.getsize(f) /1024/1024)
+    fl = f + "l"
+    scjseq.append(os.path.getsize(fl)/1024/1024)
+
+z = np.sort(np.array([nocubes, scj, scjseq]).transpose(), axis=0)
+
+compr = (z[:,1] - z[:,2]) / z[:,1]
+print(compr.min())
+print(compr.max())
+# sys.exit()
+# print(z[0])
+# print(z[:,0])
+# print(z[:,1])
+
+
+# sys.exit()
+#-- 3. plot
+fig, ax = plt.subplots()
+ax.plot(z[:,0], compr)
+# ax.plot(z[:,0], compr, marker=".")
+
+ax.set(xlabel='Number of Buildings', ylabel='file compression')
+# ax.xaxis.set_major_formatter(ticker.FormatStrFormatter('%.0f'))
+
+# ax.set_xlim(-1, 100000)
+# ax.set_ylim(0.8, 1.01)
+fig.savefig("synthetic_compression.pdf")
+plt.show()

### 2024-04-25 scripts/generate.py

@@ -25,7 +25,12 @@ def gen_vertices(total):
 
 
 def main():
-    number_of_co = 5000000
+    for t in range(3):
+        new_cityjson(random.randint(200, 1000))
+        # new_cityjson(random.randint(800000, 1200000))
+
+def new_cityjson(number_of_co):        
+    # number_of_co = 5000000
     j = {"type": "CityJSON", "version": "2.0", "CityObjects": {}, "transform": {}}
     j["transform"]["scale"] = [0.01, 0.01, 0.01]
     j["transform"]["translate"] = [85000.1, 445006.2, 0.1]
@@ -38,10 +43,9 @@ def main():
         coid = "id_" + str(i)
         j["CityObjects"][coid] = coj
     j["vertices"] = gen_vertices(number_of_co * 8) 
-
     json_str = json.dumps(j, separators=(',',':'))
-    # json_str = json.dumps(j, indent=2)
-    fname = "gen_" + str(int(number_of_co / 1000)) + "k.json"
+
+    fname = "./data/g_" + str(number_of_co) + ".json"
     with open(fname, "w") as foo:
         foo.write(json_str)
 

### 2024-04-26 paper/main.tex
Went over the paper and fixed a few things

@@
 \usepackage{booktabs}
-% \usepackage{flushend}
+\usepackage{flushend}
 \usepackage{siunitx}
@@
 
-To convert efficiently between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, freely available at \url{https://github.com/cityjson/cjseq/}.
-The program takes care of not only the geometries, but also of the materials, textures, and geometry templates that the dataset could contain.
-It contains two sub-commands: 
+To convert efficiently between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
+The command-line program takes care of converting not only the geometries, but also of the materials, textures, and geometry templates that the dataset could contain.
+It contains three sub-commands: 
 \begin{enumerate}
-  \item \texttt{cat}: based on a CityJSON file, it outputs a CityJSONSeq;
-  \item \texttt{collect}: it reads a CityJSONSeq from \emph{stdin}, collects the features, and outputs a CityJSON file (duplicates vertices are merged).
+  \item \texttt{cat}: CityJSON $\rightarrow$ CityJSONSeq;
+  \item \texttt{collect}: CityJSONSeq $\rightarrow$ CityJSON;
+  \item \texttt{filter}: to filter city objects in a CityJSONSeq, randomly or based on a bounding box.
 \end{enumerate}
 
-It should be observe that the conversion is an efficient process: the rather large dataset (\textbf{Helskinki}, which contains more than \qty{77000} buildings and whose CityJSON file is \qty{572}{\mega\byte}) takes only \qty{4.7}{\sec} to be converted to a CityJSONSeq file, and the reverse operation takes \qty{5.7}{\sec} on a standard laptop.
+It should be observe that the conversion is an efficient process: the rather large dataset (\textbf{Helskinki}, which contains more than \qty{77000} buildings and whose CityJSON file is \qty{572}{\mega\byte}), takes only \qty{4.7}{\sec} to be converted to a CityJSONSeq file, and the reverse operation takes \qty{5.7}{\sec} (on a standard laptop).
 
@@
     \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 19\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 39\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
     \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 11\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
@@
 
-With \emph{cjseq}, we have converted several publicly available files (details of files at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}), and Table~\ref{tab:stats} shows an overview of the files stored both in CityJSON and CityJSONSeq.
+With \emph{cjseq}, we have converted several publicly available files, and Table~\ref{tab:datasets} shows an overview of the files stored both in CityJSON and CityJSONSeq.
+The files themselves and their details are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}.
 
-% TODO: setup website for hosting the files
+%
 
-First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{Helsinki} it is 39\% compacter.
-The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to several millions.
+First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{Helsinki} it is 28\% compacter.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to very large integers.
 For instance, the dataset \textbf{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
@@
 
-As an experiment, we have created around 125 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated; Figure~\ref{fig:synthetic_compression} shows the results.
+As an experiment, we have created around 125 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated.
+There are no attributes, no semantics, no textures/materials. 
+Figure~\ref{fig:synthetic_compression} shows that, as the CityJSON filesize increases, the compression factor increases.
 \begin{figure}
@@
   \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
-  \caption{Compression factor of CityJSONSeq for different synthetic datasets containing only but simple cuboid buildings.}%
+  \caption{Compression factor of CityJSONSeq for different synthetic datasets containing only but simple cuboid buildings. ($\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$)}%
 \label{fig:synthetic_compression}
 \end{figure}
-Notice that as the filesize grows, the compression factor ($\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$) grows.
-The smallest file contains only but 526 buildings and its compression factor is -2\%, while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
+The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
 
@@
 
-% We can expect
+% TODO: not sure the shared vertices has a huge impact to be honest.
 
@@
 
-When a city model is stored in its entirety in one CityJSON object, we need to load the whole CityJSON object into memory in order to access the \texttt{"transform"} and \texttt{"vertices"} properties for instance.
+When a city model is stored in its entirety in one CityJSON object, we need to deserialise the whole CityJSON object into memory in order to access the \texttt{"transform"} and \texttt{"vertices"} properties for instance.
 
-With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one.
+With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one (and thus never have in memory more than the city object itself and the metadata information).
 As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
-Operations that would highly benefit from this are subsetting and merging a citymodel.
+Operations that would highly benefit from this are subsetting and merging a 3D city model.
 Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
@@
 
-We have processed all the datasets from Table~\ref{tab:datasets} with two simple Python scripts that loop through the city objects and their geometries, and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}).
+We have processed all the datasets from Table~\ref{tab:datasets} with two simple Python scripts that iterate through the city objects and their geometries, and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}) and report it at the end.
 This is just an example of a simple local operation, any other operation such as calculating the area of the façades or counting the number of windows could have been performed.
-The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script.} and the time used are shown in Table~\ref{tab:ramtime}.
+The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script} and the time used are shown in Table~\ref{tab:ramtime}.
 The scripts are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/scripts/}.
@@
 
-The results indicate that there is a very significant benefit to using CityJSONSeq, compared to regular CityJSON files, at least for operations that do not require analysing or processing city objects that are close to each other.
+The results indicate that there is a very significant benefit to using CityJSONSeq over CityJSON, at least for operations that do not require analysing or processing city objects that are close to each other.
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.

### 2024-04-26 paper/main.tex
Some fixes

@@
   \centering
-  \caption{Comparison of the processing time and max RAM use for processing the same data stored in a CityJSON file and a CityJSONSeq file.}
+  \caption{Comparison of the processing time and maximum RAM use for processing CityJSON and CityJSONSeq files.}
   \small
@@
 
-The results indicate that there is a very significant benefit to using CityJSONSeq over CityJSON, at least for operations that do not require analysing or processing city objects that are close to each other.
+Notice that the dataset \textbf{3DBV} contains the ground that has been triangulated, and a few large areas are stored as \texttt{LandUse} and contain several triangles, this is the reason why the maximum memory use is larger than for other datasets.
+
+%
+
+The results in Table~\ref{tab:ramtime} indicate that there is a very significant benefit to using CityJSONSeq over CityJSON, at least for operations that do not require analysing or processing city objects that are close to each other.
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.

### 2024-04-26 paper/main.tex
Add draft discussion

@@
 
-Notice that the dataset \textbf{3DBV} contains the ground that has been triangulated, and a few large areas are stored as \texttt{LandUse} and contain several triangles, this is the reason why the maximum memory use is larger than for other datasets.
+Notice that the dataset \textbf{3DBV} contains the ground that has been triangulated, and a few large areas are stored as \texttt{LandUse} and contain a large amount of triangles; this is the reason why the maximum memory use is larger than for other datasets.
 
@@
 
-\begin{itemize}
-  \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
-  \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
-  \item it opens up many possibilities for processing large 3dcm by using Unix pipes.
+% TODO: add conclusions
+
+While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for general applications.
+
+%
+
+CityJSONSeq, being streamable, can be processed with \emph{Unix pipes} (also called \emph{pipelines}).
+Pipelines allow us to chain several processes together, the output of a process becomes the input of the next one, and so on. 
+Given 2 processes, the 2nd one can usually start before the 1st one has finished processing all the data.
+This means in practice that the processing of a dataset can be performed by writing several programs that perform only small/atomic tasks; this makes the development and maintenance of code simpler.
+Moreover, it allows us to develop the different programs in different languages. 
+The pipeline used for preparing the datasets for this paper was actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq'), and another program written in C++ could easy be added.
+
+%
+
+The structure of CityJSONSeq is nearly the same as that of CityJSON, which means that supporting it requires a minimum amount of efforts since most of the code already existing can be used. 
+We have already added support for it in a few of the open-source CityJSON tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
+
+% We developed CityJSONSeq mainly because we wanted to be able to stream large CityJSON files from a server.
+
+% \begin{itemize}
+%   % \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
+%   % \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
+%   \item it opens up many possibilities for processing large 3dcm by using Unix pipes.
 A few of the tools developed for CityJSON are already reading/writing to stdin/stdout, so they can be chained together.
-  \item cjseqview, cjseqval are two of them, but developing others to eg filter, calculate volumes or repair geometries would be trivial
-  \item the nice part is that the tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
-  \item extensions: these should be placed in the metadata or collection
-\end{itemize}
+%   \item cjseqview, cjseqval are two of them, but developing others to eg filter, calculate volumes or repair geometries would be trivial
+%   \item the nice part is that the tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
+%   \item extensions: these should be placed in the metadata or collection
+% \end{itemize}
 
-Disadvantages:
-\begin{itemize}
-  \item  Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
-  \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.
-  \item  Appearances (esp.
+% Disadvantages:
+% \begin{itemize}
+%   \item  Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
+%   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.
+%   \item  Appearances (esp.
 Material) are relatively heavy objects with many properties.
@@
 This is just theorizing, haven't tested.
-\end{itemize}
-
-% TODO: add conclusions
+% \end{itemize}
 

### 2024-04-26 paper/main.tex
Fix many things and proof-read it
only thing not sure if shared vertices: I think they show nothing...

@@
 \usepackage{booktabs}
-\usepackage{flushend}
+% \usepackage{flushend}
 \usepackage{siunitx}
@@
 CityJSONSeq was added to the CityJSON version 2.0 standard that was recently released, and it allows us to stream very large 3D city models.
-The main idea is to decompose a CityJSON dataset into its individual features (eg each Building, each Bridge, etc.) and create several indepedent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
-We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ size, usability, etc.
-We consider CityJSONSeq to be a better format than CityJSON because: (1) once serialised it is about 10\% more compact; (2) it takes an order of magnitude less time to process; and (3) it uses significantly less memory for certain operations.
+The main idea is to decompose a CityJSON dataset into its individual city objects (eg each building, each tree, etc.) and create several independent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
+We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ filesize, usability, memory footprint, etc.
+For several use-cases, we consider CityJSONSeq to be a better format than CityJSON because: (1) once serialised it is about 10\% more compact; (2) it takes an order of magnitude less time to process; and (3) it uses significantly less memory.
 }
@@
 
-CityJSON is a JSON-based encoding for storing 3D city models that implements a subset of the latest version, v3.0.0, of the CityGML data model~\citep{OGC-CityGML3}.
-As further described in \citet{19_ogdss_cityjson}, its development started in 2017 with the aim of offering an easy-to-use and web-ready alternative to the GML-encoded CityGML files~\citep{OGC-CityGML3-XML}, which in practice can be rather verbose, difficult to parse, and complex to manipulate.
+CityJSON is a JSON-based encoding for storing 3D city models that implements a subset of version 3.0 of the CityGML data model~\citep{OGC-CityGML3}.
+As further described in \citet{19_ogdss_cityjson}, its development started in 2017 with the aim of offering an easy-to-use and web-ready alternative to the XML-encoded CityGML files~\citep{OGC-CityGML3-XML}, which in practice can be rather verbose, difficult to parse, and complex to manipulate.
 The first official release of CityJSON (version 1.0) was a success: 
 \begin{enumerate}
-  \item its JSON-based files were on average around 7X more compact than their CityGML-GML equivalents without loss of information.
-See \citet{19_ogdss_cityjson} and \url{https://www.cityjson.org/filesize/}, and also \citet{Praschl23} for a comparison with standard computer graphics formats;
+  \item its JSON-based files were on average around 7X more compact than their CityGML-XML equivalents without loss of information.
+See \citet{19_ogdss_cityjson} and \url{https://www.cityjson.org/filesize/} for details, and also \citet{Praschl23} for a comparison with standard computer graphics formats;
   \item it was adopted as an OGC Community Standard~\citep{OGC-CityJSON-v10}; 
-  \item it was adopted by the Dutch government as a 3D standard to distribute nation-wide 3D datasets; 
+  \item it was adopted by the Dutch government as a 3D standard to distribute nationwide 3D datasets; 
   \item several implementations and plugins have been developed, most notably FME\@.
@@
 
-However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries (see Figure~\ref{fig:cj_idea}) made the \emph{streaming} of very large datasets complex, if not impossible.
-Since city datasets are progressively becoming larger (it is not uncommon to have CityGML-GML files of over 2GB), the large data volume can hinder the processing of CityJSON files.
+However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries made the \emph{streaming} of very large datasets complex, if not impossible (see Figure~\ref{fig:cj_idea}).
+Since datasets of cities are progressively becoming larger (it is not uncommon to have CityGML-XML files of over 2GB), the large data volume can hinder the processing of CityJSON files.
 As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.
@@
 CityJSONSeq was added to the CityJSON version 2.0 standard that was released in 2023 (and also standardised by the OGC, see \citet{OGC-CityJSON-v20}).
-As further described in Section~\ref{sec:cityjsonseq}, the idea is to decompose a (often large) CityJSON file into its individual \emph{features} (\eg\ each \texttt{Building}, each \texttt{Bridge}, etc.) and create several JSON objects of a newly defined type \emph{CityJSONFeature}.
+As further described in Section~\ref{sec:cityjsonseq}, the idea is to decompose a (often large) CityJSON file into its individual \emph{features} (\eg\ each \texttt{Building}, each \texttt{Bridge}, etc.) and create several JSON objects of a newly defined type \texttt{CityJSONFeature}.
 These objects are either serialised into a text file, or streamed from a server to client or from one application to another.
@@
   \caption{An example of a CityJSON file.
-The vertices are stored in a global list, and the position of the vertices in that list are used in the geometries (represented by the arrows, many have been left out for clarity).}%
+The vertices are stored in a global list, and the position of the vertices in that list are used to represent the boundaries of the geometries (represented by the arrows, many have been left out for clarity).}%
 \label{fig:cj_idea}
@@
 We discuss in Section~\ref{sec:experiments} the reasons for this interesting finding.
-Finally, we will also present the open-source software for processing streams of CityJSON datasets, that is currently being developed.
 
@@
 \label{sec:cityjson}
-
-% \begin{itemize}
-%   \item global indexing of vertices make streaming impossible
-%   \item discuss the geometry templates and other metadata
-% \end{itemize}
 
@@
   \item \texttt{"vertices"}: The 3D geometric primitives are those of CityGML, and multi/composite solids with several parts are supported.
-  A geometric primitive does not list all the coordinates of its vertices, rather the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and the geometric primitives refer to the position of a vertex in that array.
+  A geometric primitive does not list all the coordinates of its vertices, instead the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and the geometric primitives refer to the position of a vertex in that array.
   The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is reused, because it has been successfully used for many years in the computer graphics community.
@@
   Third, it is very easy to convert to a representation listing all coordinates; the inverse is not true. 
-  However, this is where the streaming of geometries is problematic: the array of "vertices" contains several millions vertices, especially for very large areas, .
-  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which means waiting for millions of unused vertices to be loaded.
+  However, this list of vertices is exactly why the streaming of geometries is problematic: it can contain several millions vertices.
+  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can mean waiting for millions of unused vertices to be loaded.
   \item \texttt{"appearances"}: Both textures and materials for surfaces are supported.  
@@
   \item \texttt{"geometry-templates"} 
-  Geometry templates are used to represent identical geometries several times, they only need to be defined once (and translations/rotations/scaling are applied).
-  They are used for city objects like trees and lamp posts. 
+  Geometry templates are used to represent identical geometries several times, they only need to be defined once and translations/rotations/scaling are applied.
+  They are mostly used for city objects like trees and lamp posts. 
 \end{enumerate}
@@
 \label{sec:streaming}
-
+\begin{figure*}[th]
+  \centering
+  \includegraphics[width=0.95\linewidth]{figs/cjseq_idea}
+  \caption{The CityJSONSeq of a CityJSON dataset with two buildings contains three JSON objects: one for the metadata, plus one for each building.}%
+\label{fig:cjseq_idea}
+\end{figure*}
 In the context of geo-information, a \emph{stream} is a sequence of data that is available over a period of time, and ``can be thought of as items on a conveyor belt being processed one at a time rather than in large batches''\footnote{From \url{https://en.wikipedia.org/wiki/Stream_(computing)}}.
@@
 \emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building, for instance).
-If in the stream the linked surface has not appeared yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear.
+If in the stream the linked surface has not be seen yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear.
 It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store locally information until it appears in the stream.
-However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is copy the geometry of the 6th surface to the cube), but this means that its size is increased, and that converting back to the original file will not be possible.
+However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is copy the geometry of the 6th surface to the cube), but this means that filesize will increase, and that converting back to the original file will not be possible.
 
@@
 
-For the \emph{GeoJSON} format~\citep{IETF-GeoJSON}, which follows the Simple Feature paradigm, creating a stream is trivial since each of the features in the dataset becomes one JSON object and one line~\citep{IETF-GeoJSONSeq}.
-There are no links possible between features, they are all independent.
+For the \emph{GeoJSON} format~\citep{IETF-GeoJSON}, which follows the Simple Feature paradigm, creating a stream is trivial since each of the features in the dataset becomes one JSON object serialised to one line~\citep{IETF-GeoJSONSeq}.
+There are no links possible between features, each JSON object is independent.
 
@@
 For formats that use a global indexing of vertices (as mentioned above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
-\citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that vertices are not to be used anymore (and thus can be freed from memory).
+\citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that vertices are used anymore (and thus can be freed from memory).
 This allows us, in theory, to process/edit/manipulate infinitely large meshes, since they never have to be completely loaded in memory.
 The idea is exemplified by the construction of gridded terrains that are gigabytes in size~\citep{Isenburg06-1}.
-
 However, this cannot be implemented in CityJSON directly as all the vertices need to be in the one JSON property \texttt{"vertices"}.
@@
 
-\begin{figure*}[h]
-  \centering
-  \includegraphics[width=0.95\linewidth]{figs/cjseq_idea}
-  \caption{The CityJSONSeq of a CityJSON dataset with two buildings contains three JSON objects: one for the metadata, plus one for each building.}%
-\label{fig:cjseq_idea}
-\end{figure*}
-
-As shown in Figure~\ref{fig:cjseq_idea}, a CityJSONSeq decomposes a CityJSON object into its \emph{City Objects} (or ``features'', for instance each building, each bridge, each road, etc.), to create a sequence of several JSON objects.
-Those JSON objects are of a new type that was introduced in CityJSON v2.0: \texttt{CityJSONFeature}.
-It allows the storage of a single feature, for instance a \texttt{Building} together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}). 
-Unlike a CityJSON Object, each feature is independent. 
-It has its own list of vertices (which is thus \emph{local} to be JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
-The allowed properties are shown below; notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are many children.
+As shown in Figure~\ref{fig:cjseq_idea}, a CityJSONSeq decomposes a CityJSON object into its features to create a sequence of several JSON objects.
+Those JSON objects are of type \texttt{CityJSONFeature}, which allows the storage of a single feature, for instance a \texttt{Building} together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}).
+Unlike a CityJSON object, each feature is independent, it has its own list of vertices (which is thus \emph{local} to the JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
+The allowed properties are shown in Figure~\ref{fig:cjfeature}, notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are children.
+\begin{figure}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\caption{An example of a \texttt{CityJSONFeature} for one Building have a balcony (as its \texttt{"children"}).}%
+\label{fig:cjfeature}
+\end{figure}
 
@@
 
-To convert efficiently between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
+To convert between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
 The command-line program takes care of converting not only the geometries, but also of the materials, textures, and geometry templates that the dataset could contain.
@@
 
-It should be observe that the conversion is an efficient process: the rather large dataset (\textbf{Helskinki}, which contains more than \qty{77000} buildings and whose CityJSON file is \qty{572}{\mega\byte}), takes only \qty{4.7}{\sec} to be converted to a CityJSONSeq file, and the reverse operation takes \qty{5.7}{\sec} (on a standard laptop).
+It should be observed that the conversion is an efficient process: the rather large dataset \emph{Helskinki} from Table~\ref{tab:datasets}, which contains more than \qty{77000} buildings and whose CityJSON file is \qty{572}{\mega\byte}, takes only \qty{4.7}{\sec} to be converted to a CityJSONSeq file, and the reverse operation takes \qty{5.7}{\sec} (on a standard laptop).
 
 %%%
-\subsection{File size comparison}
+\subsection{Filesize comparison}
+
+% TODO: does the shared vertices informs us on anything? I think not and I would remove from the table.
 
@@
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 14\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
-    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 19\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
+    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
     \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 11\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 26\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
-    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 17\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
-    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 11\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
     \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
     \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
-    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 13\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
-    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 13\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
     \bottomrule
@@
       \item ${}^{\text{(a)}}$ appearance: `tex' is textures stored; `mat' is material stored
-      \item ${}^{\text{(b)}}$ compression factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$
+      \item ${}^{\text{(b)}}$ compression factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$
       \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
@@
 
-First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \textbf{Helsinki} it is 28\% compacter.
+First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\% compacter.
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to very large integers.
-For instance, the dataset \textbf{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
-The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the size of the file will grow; if they maximum vertex index is around 2000 for each feature, then the file size will be less than the original file.
+For instance, the dataset \emph{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
+The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the size of the file will grow; if the maximum vertex index is around 2000 for each feature then the filesize will be less than the original file.
 
@@
 
-Only one dataset sees its size slightly increased when serialised to a CityJSONSeq file: \textbf{Rotterdam}.
-Its CityJSONSeq size is 4\% larger than that of its CityJSON file.
+Only one dataset sees its size slightly increase by 4\% when serialised to a CityJSONSeq file: \emph{Rotterdam}.
 One reason for this is that the size of the CityJSON file is rather small in the first place, and therefore the indices are not high values.
-Having several features means that many JSON properties must be repeated for each feature, and this can cause the file size to slightly increase.
-
-%
-
 As an experiment, we have created around 125 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated.
@@
   \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
-  \caption{Compression factor of CityJSONSeq for different synthetic datasets containing only but simple cuboid buildings. ($\frac{size(CityJSON) - size(CityJSONSeq)}{CityJSON}$)}%
+  \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings. ($\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$)}%
 \label{fig:synthetic_compression}
@@
 
-% TODO: not sure the shared vertices has a huge impact to be honest.
+It should be noticed that the attributes attached to city objects, as well as the semantics attached to surfaces, have no influence on the compression factor since they are local to each city object.
+However, we can state that textures have an influence on the compression factor.
+See for instance the dataset \texttt{Helsinki} and its counterpart \texttt{Helsinki\_tex} (which is the same the same geometries and attributes, only the textures were removed from the former).
+The dataset with textures has a compression of 10\% while the one without 28\%.
+This is explained by the fact that the \texttt{"textures"} property must be used for each feature, while in a CityJSON object they are all stored at only one location.
+Since textures can be used by several features (all the bricks of a building could use the same one), this means that often the same textures are copied to several features.
 
@@
 As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
-Operations that would highly benefit from this are subsetting and merging a 3D city model.
+Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
 Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
@@
 This is just an example of a simple local operation, any other operation such as calculating the area of the façades or counting the number of windows could have been performed.
-The results for both the maximum memory footprint\footnote{The resident set size (RSS) was used, which is the portion of main memory occupied by the Python script} and the time used are shown in Table~\ref{tab:ramtime}.
+The results for both the maximum memory footprint and the time used are shown in Table~\ref{tab:ramtime}.
 The scripts are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/scripts/}.
@@
   \centering
-  \caption{Comparison of the processing time and maximum RAM use for processing CityJSON and CityJSONSeq files.}
+  \caption{Comparison of the processing time and maximum RAM use for processing CityJSON and CityJSONSeq files.
+The resident set size (RSS) is used, which is the portion of main memory occupied by the Python script.}
   \small
@@
 
-Notice that the dataset \textbf{3DBV} contains the ground that has been triangulated, and a few large areas are stored as \texttt{LandUse} and contain a large amount of triangles; this is the reason why the maximum memory use is larger than for other datasets.
+Notice that the dataset \emph{3DBV} contains not only buildings but the terrain, and a few large areas are stored as a triangulation containing a very amount of vertices; this is the reason why the maximum memory use is larger than for other datasets.
 
@@
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.
+Operations like calculating the surface of shared walls~\citep{Agugiaro22} are not designed for streams.
 
@@
 
-While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for general applications.
+While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for several use-cases.
 
@@
 
-CityJSONSeq, being streamable, can be processed with \emph{Unix pipes} (also called \emph{pipelines}).
+From a practitioners point-of-view, we should stress that CityJSONSeq can easily be processed with \emph{Unix pipes} (also called \emph{pipelines}).
 Pipelines allow us to chain several processes together, the output of a process becomes the input of the next one, and so on. 
@@
 
-The structure of CityJSONSeq is nearly the same as that of CityJSON, which means that supporting it requires a minimum amount of efforts since most of the code already existing can be used. 
+Finally, because the structure of CityJSONSeq is nearly the same as that of CityJSON, in practice adding support for CityJSONSeq requires a minimum amount of effort since most of the code already existing can be reused.
 We have already added support for it in a few of the open-source CityJSON tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
@@
 Material) are relatively heavy objects with many properties.
-If an appearance is shared across most CityObjects, duplicating it in each Feature could negate the effect of small vertex indices and actually increase the file size compared to a regular CityJSON file.
+If an appearance is shared across most CityObjects, duplicating it in each Feature could negate the effect of small vertex indices and actually increase the filesize compared to a regular CityJSON file.
 This is just theorizing, haven't tested.

### 2024-04-29 paper/main.tex
Fix 2-3 things
Clean the tex file

@@
 \author{
- Hugo Ledoux\textsuperscript{1}, Gina Stavropoulou\textsuperscript{1}, Balázs Dukai\textsuperscript{2}}
+ Hugo Ledoux\textsuperscript{1}\thanks{corresponding author}, Gina Stavropoulou\textsuperscript{1}, Balázs Dukai\textsuperscript{2}}
 
@@
 }
-
-% If the corresponding author is NOT the final author, always add a % space before the subsequent comma, i.e.
-% first author name\textsuperscript{a,}\thanks{Corresponding author} , % second author name \textsuperscript{b}, etc.
-% thanks to Niclas Borlin 05-05-2016
-% information on the corresponding author should not be used any longer and has been commented out
-% C. Heipke, Jan 03,2024
-
-% the use of the information of commissions and working groups should not be used any longer and has been commented out
-% C. Heipke, Sept. 20,2022
-%\commission{XX, }{YY} %This field is optional. If filled, XX and YY should be replaced by adequate numbers. See https://www2.isprs.org/commissions/
-%\workinggroup{XX/YY} %This field is optional.
-%\icwg{}   %This field is optional.
 
@@
   \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
-  \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings. ($\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$)}%
+  \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings.}%
 \label{fig:synthetic_compression}
@@
 Given 2 processes, the 2nd one can usually start before the 1st one has finished processing all the data.
-This means in practice that the processing of a dataset can be performed by writing several programs that perform only small/atomic tasks; this makes the development and maintenance of code simpler.
-Moreover, it allows us to develop the different programs in different languages. 
-The pipeline used for preparing the datasets for this paper was actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq'), and another program written in C++ could easy be added.
+This means that, in practice, the processing of a dataset can be performed by writing several programs that perform a few small/atomic tasks; this makes the development and maintenance of code simpler.
+Moreover, it allows us to write the code in different languages. 
+The pipeline used for preparing the datasets used in this paper was actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq'), and another program written in C++ could easy be added.
 
@@
 
-% We developed CityJSONSeq mainly because we wanted to be able to stream large CityJSON files from a server.
-
-% \begin{itemize}
-%   % \item We have kept most of the structure of CityJSON so that parsers do not need to be adapted much.
-%   % \item CityJSONSeq is an attractive alternative to CityJSON, in most uses it is better to use since it is compacter, processing it uses less memory and is faster.
-%   \item it opens up many possibilities for processing large 3dcm by using Unix pipes.
-A few of the tools developed for CityJSON are already reading/writing to stdin/stdout, so they can be chained together.
-%   \item cjseqview, cjseqval are two of them, but developing others to eg filter, calculate volumes or repair geometries would be trivial
-%   \item the nice part is that the tools do not need to be written in the same language, one pipeline can have tools built in Python and Rust and C++ and it all works nicely.
-%   \item extensions: these should be placed in the metadata or collection
-% \end{itemize}
-
-% Disadvantages:
-% \begin{itemize}
-%   \item  Software implementing CityJSON need to account for both regular CityJSON and CityJSONSeq.
-%   \item  In case of many, small, interconnecting CityObjects, a global vertex list would provide better compression than the separate vertex lists of the CityJSONFeatures.
-%   \item  Appearances (esp.
-Material) are relatively heavy objects with many properties.
-If an appearance is shared across most CityObjects, duplicating it in each Feature could negate the effect of small vertex indices and actually increase the filesize compared to a regular CityJSON file.
-This is just theorizing, haven't tested.
-% \end{itemize}
-
+\newpage
 {

### 2024-04-30 scripts/generate_adjacent.py
Add script to generate adjacent buildings (terraced houses)

@@ -1,0 +1,63 @@
+import json
+import sys
+import random
+import uuid
+
+
+
+def gen_onecube():
+    g = {"type": "Solid", "lod": "1.1", "boundaries": []}
+    b = []
+    b.append([ [0, 3, 2, 1] ])
+    b.append([ [4, 5, 6, 7] ])
+    b.append([ [0, 1, 5, 4] ])
+    b.append([ [1, 2, 6, 5] ])
+    b.append([ [2, 3, 7, 6] ])
+    b.append([ [0, 4, 7, 3] ])
+    g["boundaries"].append(b)
+    return g
+
+
+def gen_vertices(offsets):
+    vs = []
+    vs.append([offsets[0]  , offsets[1]  , offsets[2]])
+    vs.append([offsets[0]+1, offsets[1]  , offsets[2]])
+    vs.append([offsets[0]+1, offsets[1]+1, offsets[2]])
+    vs.append([offsets[0]  , offsets[1]+1, offsets[2]])
+    vs.append([offsets[0]  , offsets[1]  , offsets[2]+1])
+    vs.append([offsets[0]+1, offsets[1]  , offsets[2]+1])
+    vs.append([offsets[0]+1, offsets[1]+1, offsets[2]+1])
+    vs.append([offsets[0]  , offsets[1]+1, offsets[2]+1])
+    return vs
+
+
+def main():
+    #-- write first line
+    j = {"type": "CityJSON", "version": "2.0", "CityObjects": {}, "transform": {}, "vertices": []}
+    j["transform"]["scale"] = [1.0, 1.0, 1.0]
+    j["transform"]["translate"] = [0.0, 0.0, 0.0]
+    json_str = json.dumps(j, separators=(',',':'))
+    sys.stdout.write(json_str + "\n")
+    #-- write CityJSONFeature lines
+    for group in range(int(sys.argv[1])):
+        new_cityjson_group(int(sys.argv[2]))
+    
+
+def new_cityjson_group(number_of_co):        
+    offsets = [random.randint(0,1000), random.randint(0,1000), random.randint(0,1000)]
+    for i in range(number_of_co):
+        j = {"type": "CityJSONFeature", "CityObjects": {}, "vertices": []}
+        cgeom = gen_onecube()
+        coj = {"type": "Building", "geometry": []}
+        coj["geometry"].append(cgeom)
+        coid = str(uuid.uuid4())
+        j["id"] = coid    
+        j["CityObjects"][coid] = coj
[…]

### 2024-04-30 scripts/plot.py
Add some script file (it's now a mess, will clean)

@@ -9,7 +9,6 @@ import numpy as np
 import pandas as pd
 
 
-
 #-- 1. convert CityJSON => CityJSONSeq
 # fs = glob.glob('*.json')
 # for each in fs:
@@ -19,38 +18,50 @@ import pandas as pd
 #     os.system(cmd)
 
 
+shared = pd.read_csv("shared.csv")
+# print(shared)
+shared = shared.sort_values(by=['group'])
+# sys.exit()
+
+
 #-- 2. sizes
-fs = glob.glob('/Users/hugo/data/cityjson/CityJSONSeq-demo-files/generated/data/*.json')
-nocubes = []
+fs = glob.glob('./data/*.json')
+groups = []
 scj = []
 scjseq = []
 for f in fs:
     # print(f)
-    nocubes.append(int(f[f.find("_")+1:f.rfind(".")]))
+    groups.append(int(f[f.rfind("/")+1:f.rfind(".")]))
     scj.append(os.path.getsize(f) /1024/1024)
     fl = f + "l"
     scjseq.append(os.path.getsize(fl)/1024/1024)
 
-df = pd.DataFrame(np.array([nocubes, scj, scjseq])).transpose()
+
+df = pd.DataFrame(np.array([groups, scj, scjseq])).transpose()
 df = df.sort_values(by=[0])
+df["shared"] = shared['percentage']
+# print(df)
+# print(df.sort_values(by=[0]))
+# print(df)
+# sys.exit()
+
+
 compr = (df[1] - df[2]) / df[1]
-# sys.exit()
-# print(z[0])
-# print(z[:,0])
-# print(z[:,1])
+print(compr.min())
+print(compr.max())
 
 
-# sys.exit()
 #-- 3. plot
 fig, ax = plt.subplots()
-ax.plot(df[0], compr)
-# ax.plot(z[:,0], compr)
+# ax.plot(df[0], compr)
+ax.plot(df['shared'], compr)
 # ax.plot(z[:,0], compr, marker=".")
 
-ax.set(xlabel='Number of Buildings', ylabel='file compression')
+ax.set(xlabel='% shared vertices', ylabel='compression factor')
+# ax.set(xlabel='Number of groups', ylabel='file compression')
 # ax.xaxis.set_major_formatter(ticker.FormatStrFormatter('%.0f'))
 
 # ax.set_xlim(-1, 100000)
 # ax.set_ylim(0.8, 1.01)
[…]

### 2024-04-30 scripts/test.py

@@ -1,0 +1,7 @@
+import pandas as pd
+import numpy as np
+
+a = [1, 2, 3]
+df2 = pd.DataFrame(np.array([a, [4, 5, 6], [7, 8, 9]]), columns=['a', 'b', 'c'])
+
+print(df2)

### 2024-04-30 scripts/test2.py

@@ -1,0 +1,27 @@
+import os
+import glob
+import math
+import sys
+
+# total_buildings = 1000000
+# groups = range(0, math.floor(math.log(total_buildings, 2))+1)
+
+# for group in groups:
+#     g = round(math.pow(2, group))
+#     co = round(total_buildings / g)
+#     s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(g, co, g)
+#     # print(s)
+#     os.system(s)
+# s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(total_buildings, 1, total_buildings)
+# # print(s)
+# os.system(s)
+
+
+
+
+
+fs = glob.glob('./data/*.jsonl')
+for f in fs:
+    s = "cjseq collect -f {} > {}".format(f, f[:-1])
+    # print(s)
+    os.system(s)

### 2024-04-30 scripts/generate_adjacent.py
Updated scripts

@@ -44,7 +44,7 @@ def main():
     
 
 def new_cityjson_group(number_of_co):        
-    offsets = [random.randint(0,1000), random.randint(0,1000), random.randint(0,1000)]
+    offsets = [random.randint(0,10000), random.randint(0,10000), random.randint(0,10000)]
     for i in range(number_of_co):
         j = {"type": "CityJSONFeature", "CityObjects": {}, "vertices": []}
         cgeom = gen_onecube()

### 2024-04-30 scripts/generate_random.py

@@ -1,5 +1,6 @@
 import json
 import sys
+import uuid
 import random
 
 
@@ -25,8 +26,9 @@ def gen_vertices(total):
 
 
 def main():
-    for t in range(3):
-        new_cityjson(random.randint(200, 1000))
+    for t in range(4):
+        # new_cityjson(1000000)
+        new_cityjson(random.randint(3200000, 4000000))
         # new_cityjson(random.randint(800000, 1200000))
 
 def new_cityjson(number_of_co):        
@@ -40,12 +42,12 @@ def new_cityjson(number_of_co):
         offset += 8
         coj = {"type": "Building", "geometry": []}
         coj["geometry"].append(cgeom)
-        coid = "id_" + str(i)
+        coid = str(uuid.uuid4())
         j["CityObjects"][coid] = coj
     j["vertices"] = gen_vertices(number_of_co * 8) 
     json_str = json.dumps(j, separators=(',',':'))
 
-    fname = "./data/g_" + str(number_of_co) + ".json"
+    fname = "./data_random/g_" + str(number_of_co) + ".json"
     with open(fname, "w") as foo:
         foo.write(json_str)
 

### 2024-04-30 scripts/plot_adjacent.py

@@ -30,7 +30,7 @@ groups = []
 scj = []
 scjseq = []
 for f in fs:
-    # print(f)
+    print(f)
     groups.append(int(f[f.rfind("/")+1:f.rfind(".")]))
     scj.append(os.path.getsize(f) /1024/1024)
     fl = f + "l"
@@ -63,5 +63,5 @@ ax.set(xlabel='% shared vertices', ylabel='compression factor')
 
 # ax.set_xlim(-1, 100000)
 # ax.set_ylim(0.8, 1.01)
-# fig.savefig("synthetic_compression.pdf")
+fig.savefig("compression_adjacent.pdf")
 plt.show()

### 2024-04-30 scripts/plot_random.py

@@ -10,23 +10,22 @@ import pandas as pd
 
 
 
-#-- 1. convert CityJSON => CityJSONSeq
-# fs = glob.glob('*.json')
+# #-- 1. convert CityJSON => CityJSONSeq
+# fs = glob.glob('./data_random/*.json')
 # for each in fs:
-#     # cjseq cat -f out.json > out.jsonl
 #     cmd = "cjseq cat -f {} > {}l".format(each, each)
 #     print(cmd)
 #     os.system(cmd)
 
 
 #-- 2. sizes
-fs = glob.glob('/Users/hugo/data/cityjson/CityJSONSeq-demo-files/generated/data/*.json')
+fs = glob.glob('./data_random/*.json')
 nocubes = []
 scj = []
 scjseq = []
 for f in fs:
-    # print(f)
-    nocubes.append(int(f[f.find("_")+1:f.rfind(".")]))
+    print(f)
+    nocubes.append(int(f[f.find("g_")+2:f.rfind(".")]))
     scj.append(os.path.getsize(f) /1024/1024)
     fl = f + "l"
     scjseq.append(os.path.getsize(fl)/1024/1024)
@@ -44,13 +43,13 @@ compr = (df[1] - df[2]) / df[1]
 #-- 3. plot
 fig, ax = plt.subplots()
 ax.plot(df[0], compr)
+# ax.plot(df[0], compr, marker=".")
 # ax.plot(z[:,0], compr)
-# ax.plot(z[:,0], compr, marker=".")
 
-ax.set(xlabel='Number of Buildings', ylabel='file compression')
+ax.set(xlabel='Number of buildings', ylabel='file compression')
 # ax.xaxis.set_major_formatter(ticker.FormatStrFormatter('%.0f'))
 
 # ax.set_xlim(-1, 100000)
 # ax.set_ylim(0.8, 1.01)
-fig.savefig("synthetic_compression.pdf")
+fig.savefig("compression_random.pdf")
 plt.show()

### 2024-04-30 scripts/shared_vertices_2.py

@@ -5,8 +5,14 @@ from pathlib import Path
 import sys
 import csv
 import glob
+import matplotlib.pyplot as plt
+import matplotlib.ticker as ticker
+import matplotlib
 
-fs = glob.glob('/Users/hugo/data/cityjson/CityJSONSeq-demo-files/*.json')
+fs = glob.glob('./data/*.json')
+
+fout = open("shared.csv", 'w')
+fout.write("group,percentage")
 
 for each in fs:
     #-- json
@@ -22,7 +28,9 @@ for each in fs:
         j2 = json.loads(line)
         v_jsonl += len(j2["vertices"]) 
 
-    print(each, v_jsonl / v_json)
+    print(each, (v_jsonl - v_json) / v_json)
+    fout.write("{},{}\n".format(each, (v_jsonl - v_json) / v_json))
+
 
 
 

### 2024-04-30 scripts/size.py

@@ -4,7 +4,7 @@ import os
 import glob
 
 
-fs = glob.glob('../data/*.jsonl')
+fs = glob.glob('./data/*.jsonl')
 
 # print(fs)
 # sys.exit()

### 2024-04-30 scripts/test2.py

@@ -3,18 +3,18 @@ import glob
 import math
 import sys
 
-# total_buildings = 1000000
-# groups = range(0, math.floor(math.log(total_buildings, 2))+1)
+total_buildings = 1000000
+groups = range(0, math.floor(math.log(total_buildings, 2))+1)
 
-# for group in groups:
-#     g = round(math.pow(2, group))
-#     co = round(total_buildings / g)
-#     s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(g, co, g)
-#     # print(s)
-#     os.system(s)
-# s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(total_buildings, 1, total_buildings)
-# # print(s)
-# os.system(s)
+for group in groups:
+    g = round(math.pow(2, group))
+    co = round(total_buildings / g)
+    s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(g, co, g)
+    # print(s)
+    os.system(s)
+s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(total_buildings, 1, total_buildings)
+# print(s)
+os.system(s)
 
 
 

### 2024-04-30 paper/main.tex
Add draft section about compression_adjacent

@@
   \begin{tabular}
-    {@{}lcccccrrrcrrr@{}}\toprule
-    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{3}{c}{\textbf{vertices}}   \\ 
-    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-12} 
-     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared\footnotesize ${}^{\text{(d)}}$ \\
+    {@{}lcccccrrrcrr@{}}\toprule
+    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{2}{c}{\textbf{vertices}}   \\ 
+    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-11} 
+     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ \\
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
-    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
-    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
-    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
-    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
-    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
-    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
-    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} \\
+    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} \\
+    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} \\
+    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} \\
     \bottomrule
@@
       \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
-      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects (\eg\ the vertices of a wall shared by two buildings)
     \end{tablenotes}
@@
 There are no attributes, no semantics, no textures/materials. 
-Figure~\ref{fig:synthetic_compression} shows that, as the CityJSON filesize increases, the compression factor increases.
+Figure~\ref{fig:compression_random} shows that, as the CityJSON filesize increases, the compression factor increases.
 \begin{figure}
   \centering
-  \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
+  \includegraphics[width=\linewidth]{figs/compression_random.pdf}
   \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings.}%
-\label{fig:synthetic_compression}
+\label{fig:compression_random}
 \end{figure}
-The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
+The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of more than 12\%.
+
+%
+
+The number of shared vertices between different city objects also influences the compression factor. 
+Shared vertices are those that represent walls incident to two adjacent buildings.
+In CityJSON they are conceptually the same vertices and each of the surfaces refer to them, but in CityJSONSeq they have to be listed separately in each of the buildings.
+It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%).
+To understand the correlation between the compression factor of CityJSONSeq and the percentage of shared vertices in a datasets, we have generated several datasets also containing cuboid buildings. 
+However, the distribution of the buildings is not random: we enforced that several buildings are adjacent to others (so that they share vertices with other buildings).
+The relationship between the compression and the percentage of shared vertices can be seen in Figure~\ref{fig:compression_adjacent} for around 100 datasets containing exactly \num{1000000} buildings.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figs/compression_adjacent.pdf}
+  \caption{Compression factor of CityJSONSeq files with respect to the percentage of shared vertices in a dataset that was synthetically generated.}% 
+\label{fig:compression_adjacent}
+\end{figure}
+If the number of shared vertices is 0\%, this means that we have \num{1000000} buildings that are disconnected; in this case we obtain a compression factor of around 8\%.
+If all the buildings are adjacent to another one (shared vertices are nearly 100\%), then we can see that the compression factor is nearly -10\% (which means that the size of the CityJSONSeq file is larger than that of the CityJSON file).
 
@@
 
-\newpage
+% \newpage
 {

### 2024-05-01 paper/main.tex
Some proof-reading this morning

@@
 
-We present \emph{CityJSON Text Sequences} (CityJSONSeq), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
-CityJSONSeq was added to the CityJSON version 2.0 standard that was recently released, and it allows us to stream very large 3D city models.
+We present \emph{CityJSON Text Sequences} (CityJSONSeq in short), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
+CityJSONSeq was added to the CityJSON version 2.0 standard to allow us to stream very large 3D city models.
 The main idea is to decompose a CityJSON dataset into its individual city objects (eg each building, each tree, etc.) and create several independent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
@@
 
-\keywords{3D city modelling, CityJSON, CityGML, streaming}
+\keywords{3D city modelling, CityJSON, streaming, massive datasets, CityGML}
 \maketitle
@@
 
-CityJSON is a JSON-based encoding for storing 3D city models that implements a subset of version 3.0 of the CityGML data model~\citep{OGC-CityGML3}.
+CityJSON is a JSON-based encoding for storing 3D city models that implements a subset of the CityGML data model version 3.0~\citep{OGC-CityGML3}.
 As further described in \citet{19_ogdss_cityjson}, its development started in 2017 with the aim of offering an easy-to-use and web-ready alternative to the XML-encoded CityGML files~\citep{OGC-CityGML3-XML}, which in practice can be rather verbose, difficult to parse, and complex to manipulate.
@@
 \begin{enumerate}
-  \item its JSON-based files were on average around 7X more compact than their CityGML-XML equivalents without loss of information.
+  \item its JSON-based files were on average around 7 times more compact than their CityGML-XML equivalents without loss of information.
 See \citet{19_ogdss_cityjson} and \url{https://www.cityjson.org/filesize/} for details, and also \citet{Praschl23} for a comparison with standard computer graphics formats;
@@
 However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries made the \emph{streaming} of very large datasets complex, if not impossible (see Figure~\ref{fig:cj_idea}).
-Since datasets of cities are progressively becoming larger (it is not uncommon to have CityGML-XML files of over 2GB), the large data volume can hinder the processing of CityJSON files.
 As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.
+Since datasets of 3D cities are nowadays very large (it is not uncommon to have CityGML-XML files over 2GB), the processing CityJSON files is an issue in practice.
 
@@
 These objects are either serialised into a text file, or streamed from a server to client or from one application to another.
-This allows us to avoid a long list of vertices that need to be indexed, as it is the case with CityJSON files (see Section~\ref{sec:cityjson} for details).
+This allows us to avoid a large list of vertices that need to be indexed, as it is the case with CityJSON files (see Section~\ref{sec:cityjson} for details).
 \begin{figure}
   \centering
-  \includegraphics[width=0.95\linewidth]{figs/cj_idea}
+  \includegraphics[width=\linewidth]{figs/cj_idea}
   \caption{An example of a CityJSON file.
@@
 
-We have implemented and released an open-source software to convert between CityJSON and CityJSONSeq objects; we describe it in Section~\ref{sec:experiments}.
+In Section~\ref{sec:experiments}, we describe  the open-source software we have developed to convert between CityJSON and CityJSONSeq files.
 We also analyse the size of CityJSONSeq files versus CityJSON files, and that for several real-world datasets.
@@
   \item \texttt{"version"}: \texttt{"2.0"} is the current version;
-  \item \texttt{"metadata"}: different metadata related to the dataset can be stored. The most important is the definition of the coordinate reference system (CRS)
+  \item \texttt{"metadata"}: different metadata related to the dataset can be stored. The most important is the definition of the coordinate reference system (CRS).
   \item \texttt{"transform"}: CityJSON \texttt{"vertices"} are compressed and stored as integers only.
@@
   \item \texttt{"CityObjects"}: a dictionary where the properties are the identifiers of the city objects (\emph{IDs}), which are any CityGML city object (for instance a \texttt{Building}, a \texttt{BuildingPart}, a \texttt{SolitaryVegetationObject}, etc.).
-  Each of the city objects is listed one after the other, even if some are \texttt{"children"} of others.
+  The city objects are listed one after the other, even if some are \texttt{"children"} of others.
   As an example, for a \texttt{Building} containing 2 parts, the 3 objects will be represented at the same level and linked by their \emph{IDs}, as shown in Figure~\ref{fig:parents_children}.
@@
   The fact that a dictionary is used means that developers have direct access to the city objects through their IDs (and also in constant time if a hashmap is used to implement the dictionary while parsing the file).
-  \item \texttt{"vertices"}: The 3D geometric primitives are those of CityGML, and multi/composite solids with several parts are supported.
+  \item \texttt{"vertices"}: The 3D geometric primitives in CityJSON are those of the CityGML data model, which means that multi/composite solids with several parts and/or cavities are supported.
   A geometric primitive does not list all the coordinates of its vertices, instead the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and the geometric primitives refer to the position of a vertex in that array.
-  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is reused, because it has been successfully used for many years in the computer graphics community.
+  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is thus reused, because it has been successfully used for many years in the computer graphics community.
   There are several advantages to this approach.
@@
   Third, it is very easy to convert to a representation listing all coordinates; the inverse is not true. 
-  However, this list of vertices is exactly why the streaming of geometries is problematic: it can contain several millions vertices.
-  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can mean waiting for millions of unused vertices to be loaded.
+  However, this list of vertices is the reason why the streaming of geometries is problematic, since in practice it can contain several millions vertices.
+  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can mean waiting for millions of unused vertices to be deserialised.
   \item \texttt{"appearances"}: Both textures and materials for surfaces are supported.  
@@
   \item \texttt{"geometry-templates"} 
-  Geometry templates are used to represent identical geometries several times, they only need to be defined once and translations/rotations/scaling are applied.
-  They are mostly used for city objects like trees and lamp posts. 
+  Geometry templates are geometries defined once and reused by applying a translation, a rotation, and/or a scaling.
+  They are mostly used for city objects like trees, bus stops, and lamp posts. 
 \end{enumerate}
@@
 For GML-based formats (which are feature-centric), modifying a file for the purpose of streaming is usually a simple task that involves sending the features in the dataset one-by-one.
-However, notice that this is only true for simple files and GML features stored following the Simple Feature paradigm~\citep{OGC-SF}.
-For more complex data models like CityGML where \emph{Implicit Geometries} and \emph{XLinks} are used, the conversion often needs a large amount of pre-processing.
-\emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building, for instance).
-If in the stream the linked surface has not be seen yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear.
-It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store locally information until it appears in the stream.
-However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is copy the geometry of the 6th surface to the cube), but this means that filesize will increase, and that converting back to the original file will not be possible.
+However, notice that this is only true for GML files that follow the Simple Feature paradigm~\citep{OGC-SF}.
+For more complex data models like CityGML where geometry templates and \emph{XLinks} are used, streaming often needs a large amount of pre-processing.
+\emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building for instance).
+If in the stream the linked surface has not be seen yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear (\eg\ to calculate its volume).
+It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store extra information until it appears in the stream.
+However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is, in our example, copy the geometry of the 6th surface to the cube), but this means that filesize will increase, and that converting back to the original file will not be possible.
 
@@
 For formats that use a global indexing of vertices (as mentioned above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
-\citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that vertices are used anymore (and thus can be freed from memory).
+\citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that specific vertices are not used anymore in the stream (and thus can be freed from memory).
 This allows us, in theory, to process/edit/manipulate infinitely large meshes, since they never have to be completely loaded in memory.
 The idea is exemplified by the construction of gridded terrains that are gigabytes in size~\citep{Isenburg06-1}.
-However, this cannot be implemented in CityJSON directly as all the vertices need to be in the one JSON property \texttt{"vertices"}.
+However, this cannot be implemented in CityJSON directly because all the vertices need to be listed in the JSON property \texttt{"vertices"}.
 
@@
 Those JSON objects are of type \texttt{CityJSONFeature}, which allows the storage of a single feature, for instance a \texttt{Building} together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}).
-Unlike a CityJSON object, each feature is independent, it has its own list of vertices (which is thus \emph{local} to the JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
+Each feature is independent, it has its own list of vertices (which is thus \emph{local} to the JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
 The allowed properties are shown in Figure~\ref{fig:cjfeature}, notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are children.
@@
   \item the first JSON Object must be of type \texttt{CityJSON};
-  \item the following JSON Objects are of type \texttt{CityJSONFeature}.
+  \item the following JSON Objects must be of type \texttt{CityJSONFeature}.
 \end{enumerate}
 
-Note that a \texttt{CityJSONFeature} object does not contain all the information that is required for parsing the feature. 
-Most commonly, the \texttt{"transform"} property, the CRS, and the \texttt{"geometry-templates"} must be known by the client in order to correctly reconstruct and georeference the city objects. 
+Note that a \texttt{CityJSONFeature} object does not contain all the information that is required for reconstructing the feature. 
+Most commonly, the \texttt{"transform"} property, the CRS, and the \texttt{"geometry-templates"} must be known in order to correctly reconstruct and georeference the city objects. 
 The rule \#4 ensures that those are available. 
@@
 
-The CityJSONSeq for Figure~\ref{fig:cjseq_idea} is shown in Figure~\ref{fig:stream}:
+The CityJSONSeq for Figure~\ref{fig:cjseq_idea} is shown in Figure~\ref{fig:stream}.
 \begin{figure*}
@@
 
-To convert between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
-The command-line program takes care of converting not only the geometries, but also of the materials, textures, and geometry templates that the dataset could contain.
+To convert between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, which is available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
+The command-line program takes care of converting not only the geometries, but also the materials, the textures, and the geometry templates that the dataset could contain.
 It contains three sub-commands: 
@@
 \subsection{Filesize comparison}
-
-% TODO: does the shared vertices informs us on anything? I think not and I would remove from the table.
 

### 2024-05-01 paper/main.tex
Some more

@@
 
-We present \emph{CityJSON Text Sequences} (CityJSONSeq in short), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
+We introduce \emph{CityJSON Text Sequences} (CityJSONSeq in short), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
 CityJSONSeq was added to the CityJSON version 2.0 standard to allow us to stream very large 3D city models.
-The main idea is to decompose a CityJSON dataset into its individual city objects (eg each building, each tree, etc.) and create several independent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
+The main idea is to decompose a CityJSON dataset into its individual city objects (each building, each tree, etc.) and create several independent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
 We elaborate on the engineering decisions that were taken to develop CityJSONSeq, we present the open-source software we have developed to convert to and from CityJSONSeq, and we discuss different aspects of the new format, \eg\ filesize, usability, memory footprint, etc.
@@
 
-However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries made the \emph{streaming} of very large datasets complex, if not impossible (see Figure~\ref{fig:cj_idea}).
+However, the version 1.0 of CityJSON had one limitation: its structure for storing the coordinates of the geometries (see Figure~\ref{fig:cj_idea}) made the \emph{streaming} of very large datasets complex, if not impossible.
 As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.
-Since datasets of 3D cities are nowadays very large (it is not uncommon to have CityGML-XML files over 2GB), the processing CityJSON files is an issue in practice.
+Since datasets of 3D cities are nowadays very large (it is not uncommon to have CityGML-XML files over 2GB), the processing CityJSON files is in practice an issue.
 

### 2024-05-01 paper/main.tex
Add storage paragraph

@@
 
+Futhermore, the CityJSON specification does not prescribe the storage of CityJSONSeq, only the structure of a CityJSONSeq stream (see the five constraints above).
+In practice, CityJSONSeq can be stored in a variety of ways, eg in a single file, each feature in a separate file, in a database etc., and the optimal storage solution depends on the implementing application.
+
 %

### 2024-05-01 paper/main.tex
Be specific about what to alter, because we might not have a file at all.

@@
 Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
-Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
+Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the CityJSON object.
 

### 2024-05-01 paper/main.tex
Some sentence fixed

--- 2024-05-01 paper/main.tex
@@
 In Section~\ref{sec:experiments}, we describe  the open-source software we have developed to convert between CityJSON and CityJSONSeq files.
-We also analyse the size of CityJSONSeq files versus CityJSON files, and that for several real-world datasets.
-It can be observed that one advantage of CityJSONSeq, besides that files containing thousands of features can be processed with a low memory footprint, is that it compresses further the CityJSON files by around 12\%, sometimes more.
+We also analyse the filesizes of CityJSON and CityJSONSeq for several real-world datasets and synthetic datasets we built.
+It can be observed that one advantage of CityJSONSeq, besides that files containing several thousands of features can be streamed, is that it compresses further the CityJSON files by around 12\%, sometimes more.
 We discuss in Section~\ref{sec:experiments} the reasons for this interesting finding.

### 2024-05-01 paper/main.tex
Put back the shared vertices for hte datasets

@@
   \begin{tabular}
-    {@{}lcccccrrrcrr@{}}\toprule
-    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{2}{c}{\textbf{vertices}}   \\ 
-    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-11} 
-     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ \\
+    {@{}lcccccrrrcrrr@{}}\toprule
+    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{3}{c}{\textbf{vertices}}   \\ 
+    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-12} 
+     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared\footnotesize ${}^{\text{(d)}}$ \\
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} \\
-    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} \\
-    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} \\
-    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} \\
-    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} \\
-    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} \\
-    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} \\
-    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
+    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
+    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
+    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
     \bottomrule
@@
       \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
+      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects (\eg\ the vertices of a wall shared by two buildings)
     \end{tablenotes}
@@
 There are no attributes, no semantics, no textures/materials. 
-Figure~\ref{fig:compression_random} shows that, as the CityJSON filesize increases, the compression factor increases.
+Figure~\ref{fig:synthetic_compression} shows that, as the CityJSON filesize increases, the compression factor increases.
 \begin{figure}
   \centering
-  \includegraphics[width=\linewidth]{figs/compression_random.pdf}
+  \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
   \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings.}%
-\label{fig:compression_random}
+\label{fig:synthetic_compression}
 \end{figure}
-The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of more than 12\%.
-
-%
-
-The number of shared vertices between different city objects also influences the compression factor. 
-Shared vertices are those that represent walls incident to two adjacent buildings.
-In CityJSON they are conceptually the same vertices and each of the surfaces refer to them, but in CityJSONSeq they have to be listed separately in each of the buildings.
-It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%).
-To understand the correlation between the compression factor of CityJSONSeq and the percentage of shared vertices in a datasets, we have generated several datasets also containing cuboid buildings. 
-However, the distribution of the buildings is not random: we enforced that several buildings are adjacent to others (so that they share vertices with other buildings).
-The relationship between the compression and the percentage of shared vertices can be seen in Figure~\ref{fig:compression_adjacent} for around 100 datasets containing exactly \num{1000000} buildings.
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figs/compression_adjacent.pdf}
-  \caption{Compression factor of CityJSONSeq files with respect to the percentage of shared vertices in a dataset that was synthetically generated.}% 
-\label{fig:compression_adjacent}
-\end{figure}
-If the number of shared vertices is 0\%, this means that we have \num{1000000} buildings that are disconnected; in this case we obtain a compression factor of around 8\%.
-If all the buildings are adjacent to another one (shared vertices are nearly 100\%), then we can see that the compression factor is nearly -10\% (which means that the size of the CityJSONSeq file is larger than that of the CityJSON file).
+The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
 
@@
 
-% \newpage
+\newpage
 {

### 2024-05-01 paper/main.tex
Proofreading 1

@@
 
-With \emph{cjseq}, we have converted several publicly available files, and Table~\ref{tab:datasets} shows an overview of the files stored both in CityJSON and CityJSONSeq.
-The files themselves and their details are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}.
+We have converted with \emph{cjseq} several publicly available files, and Table~\ref{tab:datasets} shows an overview of the files stored both in CityJSON and CityJSONSeq.
+The files themselves are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}.
 
@@
 
-First observe that---contrary to intuition---the size of a dataset serialised as a CityJSONSeq file is around 12\% more compact than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\% compacter.
-The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0''), and do not increase to very large integers.
+First observe that---contrary to intuition---the filesize of a dataset serialised as a CityJSONSeq file is around 12\% compacter than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\%.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and do not increase to very large integers.
 For instance, the dataset \emph{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
-The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the size of the file will grow; if the maximum vertex index is around 2000 for each feature then the filesize will be less than the original file.
+The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the filesize will grow; if the maximum vertex index is around 2000 for each feature then the filesize will be reduced.
 
@@
 
-Only one dataset sees its size slightly increase by 4\% when serialised to a CityJSONSeq file: \emph{Rotterdam}.
-One reason for this is that the size of the CityJSON file is rather small in the first place, and therefore the indices are not high values.
-As an experiment, we have created around 125 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated.
-There are no attributes, no semantics, no textures/materials. 
+Only one dataset sees its filesize slightly increase, by 4\%, when serialised to a CityJSONSeq file: \emph{Rotterdam}.
+One reason for this is that the size of the CityJSON file is rather small in the first place, and therefore the indices are not large integers.
+As an experiment, we have created around 100 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated.
+There are no attributes, no semantics, and no textures/materials. 
 Figure~\ref{fig:synthetic_compression} shows that, as the CityJSON filesize increases, the compression factor increases.
@@
 \end{figure}
-The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
+The smallest file contains 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
 

### 2024-05-01 paper/main.tex
Finish my proofreading

@@
 There are no attributes, no semantics, and no textures/materials. 
-Figure~\ref{fig:synthetic_compression} shows that, as the CityJSON filesize increases, the compression factor increases.
+Figure~\ref{fig:compression_random.pdfn} shows that, as the CityJSON filesize increases, the compression factor increases.
 \begin{figure}
   \centering
-  \includegraphics[width=\linewidth]{figs/synthetic_compression.pdf}
+  \includegraphics[width=\linewidth]{figs/compression_random.pdf}
   \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings.}%
-\label{fig:synthetic_compression}
+\label{fig:compression_random}
 \end{figure}
-The smallest file contains 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of 18\%.
+The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of more than 12\%.
 
@@
 
-It should be noticed that the attributes attached to city objects, as well as the semantics attached to surfaces, have no influence on the compression factor since they are local to each city object.
+The number of shared vertices between different city objects also influences the compression factor. 
+Shared vertices are those used to represent walls incident to two adjacent buildings.
+In CityJSON they are conceptually the same vertices and each of the surfaces refer to them, but in CityJSONSeq they have to be listed separately in each of the buildings.
+It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%).
+To understand the correlation between the compression factor and the percentage of shared vertices in a datasets, we have modified the script to generate random cuboid buildings: the distribution of the buildings is not random, we have enforced that several buildings are adjacent to others (so that they share vertices with other buildings).
+The relationship between the compression and the percentage of shared vertices can be seen in Figure~\ref{fig:compression_adjacent} for around 100 datasets containing exactly \num{1000000} buildings.
+\begin{figure}
+  \centering
+  \includegraphics[width=\linewidth]{figs/compression_adjacent.pdf}
+  \caption{Compression factor of CityJSONSeq files with respect to the percentage of shared vertices in a dataset that was synthetically generated.}% 
+\label{fig:compression_adjacent}
+\end{figure}
+If the number of shared vertices is 0\% this means that we have \num{1000000} buildings that are disconnected; in this case we obtain a compression factor of around 8\% (as was the case in Figure~\ref{fig:compression_random}).
+If all the buildings are adjacent to another one (thus nearly 100\% of the vertices are shared), then we can see that the compression factor is about -10\% (which means that the size of the CityJSONSeq file is larger than that of the CityJSON file).
+%
+
+It should also be noticed that the attributes attached to city objects, as well as the semantics attached to surfaces, have no influence on the compression factor since they are local to each city object.
 However, we can state that textures have an influence on the compression factor.
-See for instance the dataset \texttt{Helsinki} and its counterpart \texttt{Helsinki\_tex} (which is the same the same geometries and attributes, only the textures were removed from the former).
+See for instance the dataset \emph{Helsinki} and its counterpart \emph{Helsinki\_tex} (which is the same the same geometries and attributes, only the textures were removed).
 The dataset with textures has a compression of 10\% while the one without 28\%.
 This is explained by the fact that the \texttt{"textures"} property must be used for each feature, while in a CityJSON object they are all stored at only one location.
-Since textures can be used by several features (all the bricks of a building could use the same one), this means that often the same textures are copied to several features.
+Since textures can be used by several features (all the bricks of a building could use the same one), this means that often the same properties for textures are copied to several features.
 
@@
 We compare the speed and memory footprint of accessing each city object in a dataset.
-This operation is common in applications that manipulate city models.
+This operation is common in applications that manipulate 3D city models.
 
@@
 
-With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one (and thus never have in memory more than the city object itself and the metadata information).
+With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one (and thus never have in memory more than the city object itself and the first JSON object in the stream).
 As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
-Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
-Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
+% Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
 
@@
 
-Notice that the dataset \emph{3DBV} contains not only buildings but the terrain, and a few large areas are stored as a triangulation containing a very amount of vertices; this is the reason why the maximum memory use is larger than for other datasets.
+Notice that the dataset \emph{3DBV} contains not only buildings but the terrain, and a few large areas are stored as a triangulation containing a very large amount of vertices; this is the reason why the maximum memory use is larger than for other datasets.
 
@@
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.
-Operations like calculating the surface of shared walls~\citep{Agugiaro22} are not designed for streams.
+Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
+Operations like calculating the surface of shared walls~\citep{Agugiaro22} are however not designed for streams.
 
@@
 Moreover, it allows us to write the code in different languages. 
-The pipeline used for preparing the datasets used in this paper was actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq'), and another program written in C++ could easy be added.
+The pipelines used for preparing the datasets used in this paper were actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq' and `cjval'), and another program written in C++ could easy be added.
 

### 2024-05-02 paper/main.tex
Spell-check

@@
 
-Futhermore, the CityJSON specification does not prescribe the storage of CityJSONSeq, only the structure of a CityJSONSeq stream (see the five constraints above).
+Furthermore, the CityJSON specification does not prescribe the storage of CityJSONSeq, only the structure of a CityJSONSeq stream (see the five constraints above).
 In practice, CityJSONSeq can be stored in a variety of ways, eg in a single file, each feature in a separate file, in a database etc., and the optimal storage solution depends on the implementing application.
@@
       \item ${}^{\text{(a)}}$ appearance: `tex' is textures stored; `mat' is material stored
-      \item ${}^{\text{(b)}}$ compression factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$
+      \item ${}^{\text{(b)}}$ compressi{}on factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$
       \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
-      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects (\eg\ the vertices of a wall shared by two buildings)
+      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects
     \end{tablenotes}
@@
 In CityJSON they are conceptually the same vertices and each of the surfaces refer to them, but in CityJSONSeq they have to be listed separately in each of the buildings.
-It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%).
+It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%, except 2 datasets have around 20\%).
 To understand the correlation between the compression factor and the percentage of shared vertices in a datasets, we have modified the script to generate random cuboid buildings: the distribution of the buildings is not random, we have enforced that several buildings are adjacent to others (so that they share vertices with other buildings).
@@
 Moreover, it allows us to write the code in different languages. 
-The pipelines used for preparing the datasets used in this paper were actually a mix of Python (the program `cjio' and `cjdb') and Rust (`cjseq' and `cjval'), and another program written in C++ could easy be added.
+The pipelines used for preparing the datasets used in this paper were actually a mix of Python (the program \emph{cjio} and \emph{cjdb}) and Rust (\emph{cjseq} and \emph{cjval}), and another program written in C++ could easy be added.
 

### 2024-05-02 scripts/generate_adjacent.py
Clean scripts to generate data and test them and extract statistics

@@ -2,7 +2,37 @@ import json
 import sys
 import random
 import uuid
+import os
+import glob
+import math
 
+def main():
+    total_buildings = 100
+    groups = range(0, math.floor(math.log(total_buildings, 2))+1)
+    for group in groups:
+        g = round(math.pow(2, group))
+        co = round(total_buildings / g)
+        write_cjseq(g, co)
+    write_cjseq(1, total_buildings)
+    #-- convert to cityjson
+    fs = glob.glob('./data_adjacent/*.jsonl')
+    for f in fs:
+        s = "cjseq collect -f {} > {}".format(f, f[:-1])
+        # print(s)
+        os.system(s)
+
+def write_cjseq(no_groups, no_co):
+    fname = "./data_adjacent/" + str(no_co) + ".jsonl"
+    fo = open(fname, "w")
+    #-- write first line
+    j = {"type": "CityJSON", "version": "2.0", "CityObjects": {}, "transform": {}, "vertices": []}
+    j["transform"]["scale"] = [1.0, 1.0, 1.0]
+    j["transform"]["translate"] = [0.0, 0.0, 0.0]
+    json_str = json.dumps(j, separators=(',',':'))
+    fo.write(json_str + "\n")
+    #-- write CityJSONFeature lines
+    for group in range(no_groups):
+        new_cityjson_group(no_co, fo)
 
 
 def gen_onecube():
@@ -31,19 +61,9 @@ def gen_vertices(offsets):
     return vs
 
 
-def main():
-    #-- write first line
-    j = {"type": "CityJSON", "version": "2.0", "CityObjects": {}, "transform": {}, "vertices": []}
-    j["transform"]["scale"] = [1.0, 1.0, 1.0]
-    j["transform"]["translate"] = [0.0, 0.0, 0.0]
-    json_str = json.dumps(j, separators=(',',':'))
-    sys.stdout.write(json_str + "\n")
-    #-- write CityJSONFeature lines
-    for group in range(int(sys.argv[1])):
-        new_cityjson_group(int(sys.argv[2]))
     
 
-def new_cityjson_group(number_of_co):        
+def new_cityjson_group(number_of_co, fo):        
     offsets = [random.randint(0,10000), random.randint(0,10000), random.randint(0,10000)]
     for i in range(number_of_co):
         j = {"type": "CityJSONFeature", "CityObjects": {}, "vertices": []}
[…]

### 2024-05-02 scripts/shared_vertices.py

@@ -1,47 +1,36 @@
-"""
-Script for counting how many CityObjects share a shared vertex.
-It writes a CSV file with the shared vertex count for each vertex, to the same
-directory as the input file.
-"""
+
 import json
 import sys
 from pathlib import Path
 import sys
 import csv
+import glob
+import matplotlib.pyplot as plt
+import matplotlib.ticker as ticker
+import matplotlib
 
-if __name__ == '__main__':
-    inpath = Path(sys.argv[1]).resolve()
+fs = glob.glob('./data/*.json')
 
-    with inpath.open("r") as f:
-        data = json.load(f)
+fout = open("shared.csv", 'w')
+fout.write("group,percentage")
 
-    vertex_counts = {i: set() for i in range(len(data['vertices']))}
+for each in fs:
+    #-- json
+    with Path(each).resolve().open("r") as f:
+        j = json.load(f)
+    v_json = len(j["vertices"])
 
-    for co_id, co in data["CityObjects"].items():
-        for geometry in co["geometry"]:
-            for surface in geometry["boundaries"]:
-                for ring in surface:
-                    for vertex in ring:
-                        vertex_counts[vertex].add(co_id)
+    #-- jsonl
+    eachl = each + "l"
+    f = open(eachl)
+    v_jsonl = 0
+    for (i, line) in enumerate(f.readlines()):
+        j2 = json.loads(line)
+        v_jsonl += len(j2["vertices"]) 
 
-    vertex_counts_sorted = dict(sorted(vertex_counts.items(), key=lambda item: len(item[1]), reverse=True))
+    print(each, (v_jsonl - v_json) / v_json)
+    fout.write("{},{}\n".format(each, (v_jsonl - v_json) / v_json))
 
-    nr_shared = 0
-    size_vertex = sys.getsizeof(data['vertices'][0])
-    total_size_vertices = size_vertex * len(data['vertices'])
-    total_size_vertices_unshared = 0
-    with inpath.parent.joinpath(inpath.stem).with_suffix(".csv").open("w") as f:
-        csv_writer = csv.writer(f)
-        csv_writer.writerow(["Vertex", "Count"])
-        for vertex, ids in vertex_counts_sorted.items():
-            if len(ids) > 1:
-                nr_shared += 1
-                total_size_vertices_unshared += size_vertex * len(ids)
-                csv_writer.writerow([vertex, len(ids)])
-            else:
[…]

### 2024-05-02 scripts/shared_vertices_2.py

@@ -1,36 +1,0 @@
-
-import json
-import sys
-from pathlib import Path
-import sys
-import csv
-import glob
-import matplotlib.pyplot as plt
-import matplotlib.ticker as ticker
-import matplotlib
-
-fs = glob.glob('./data/*.json')
-
-fout = open("shared.csv", 'w')
-fout.write("group,percentage")
-
-for each in fs:
-    #-- json
-    with Path(each).resolve().open("r") as f:
-        j = json.load(f)
-    v_json = len(j["vertices"])
-
-    #-- jsonl
-    eachl = each + "l"
-    f = open(eachl)
-    v_jsonl = 0
-    for (i, line) in enumerate(f.readlines()):
-        j2 = json.loads(line)
-        v_jsonl += len(j2["vertices"]) 
-
-    print(each, (v_jsonl - v_json) / v_json)
-    fout.write("{},{}\n".format(each, (v_jsonl - v_json) / v_json))
-
-
-
-

### 2024-05-02 scripts/test2.py

@@ -1,27 +1,0 @@
-import os
-import glob
-import math
-import sys
-
-total_buildings = 1000000
-groups = range(0, math.floor(math.log(total_buildings, 2))+1)
-
-for group in groups:
-    g = round(math.pow(2, group))
-    co = round(total_buildings / g)
-    s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(g, co, g)
-    # print(s)
-    os.system(s)
-s = "python generate_adjacent.py {} {} > ./data/{}.jsonl".format(total_buildings, 1, total_buildings)
-# print(s)
-os.system(s)
-
-
-
-
-
-fs = glob.glob('./data/*.jsonl')
-for f in fs:
-    s = "cjseq collect -f {} > {}".format(f, f[:-1])
-    # print(s)
-    os.system(s)

### 2024-05-02 paper/main.tex
introductions

@@
 As further defined in Section~\ref{sec:streaming}, streaming refers to the possibility of downloading/transferring/processing a dataset without having to load it all in memory.
-Since datasets of 3D cities are nowadays very large (it is not uncommon to have CityGML-XML files over 2GB), the processing CityJSON files is in practice an issue.
+Given the increasing size of datasets of 3D cities, with CityGML-XML files often exceeding 2GB, the processing of CityJSON files has become a practical challenge.
 
@@
 
-We present in this paper \emph{CityJSON Text Sequences} (\emph{CityJSONSeq} in short in the following), a format based on \emph{JSON Text Sequences}~\citep{IETF-JSONSeq} and CityJSON, and inspired by \emph{GeoJSON Text Sequences}~\citep{IETF-GeoJSONSeq}.
+We present in this paper \emph{CityJSON Text Sequences} (henceforth referred to as \emph{CityJSONSeq}), a format based on \emph{JSON Text Sequences}~\citep{IETF-JSONSeq} and CityJSON, and inspired by \emph{GeoJSON Text Sequences}~\citep{IETF-GeoJSONSeq}.
 CityJSONSeq was added to the CityJSON version 2.0 standard that was released in 2023 (and also standardised by the OGC, see \citet{OGC-CityJSON-v20}).

### 2024-05-02 paper/main.tex
structure

@@
   A geometric primitive does not list all the coordinates of its vertices, instead the coordinates of the vertices are stored in a separate array (the \texttt{"vertices"} property of the CityJSON object), and the geometric primitives refer to the position of a vertex in that array.
-  The indexing mechanism of the format \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}} is thus reused, because it has been successfully used for many years in the computer graphics community.
+  This indexing mechanism has been successfully used for many years by the computer graphics community in formats as \emph{Wavefront OBJ}\footnote{\url{https://en.wikipedia.org/wiki/Wavefront_.obj_file}}.
   There are several advantages to this approach.
   First, the files can be compressed: 3D vertices are often shared by several surfaces, and repeating them can be costly, especially if they are very precise (sub-millimetre precision is often used).
-  Second, this approach increases the topological relationships that are explicitly stored in the file, and several operations (\eg\ are 2 buildings adjacent?) can be sped up and made more robust.
-  Third, it is very easy to convert to a representation listing all coordinates; the inverse is not true. 
+  Second, this approach increases the topological relationships that are explicitly stored in the file, and several operations (\eg\ determining building adjacency) can be sped up and made more robust.
+  Third, it is very easy to convert all coordinates to a representation listing; the inverse is not true. 
   However, this list of vertices is the reason why the streaming of geometries is problematic, since in practice it can contain several millions vertices.
-  To be able to reconstruct one Building, all the \texttt{"vertices"} need to be in memory, which can mean waiting for millions of unused vertices to be deserialised.
+  To be able to reconstruct a single Building, all the \texttt{"vertices"} need to loaded in memory, which can mean waiting for millions of unused vertices to be deserialised.
   \item \texttt{"appearances"}: Both textures and materials for surfaces are supported.  
   The material of a surface is represented with the X3D specifications\footnote{\url{https://en.wikipedia.org/wiki/X3D}}, and for the textures the COLLADA specifications\footnote{\url{https://www.khronos.org/collada/}} are reused.
-  \item \texttt{"geometry-templates"} 
+  \item \texttt{"geometry-templates"}: 
   Geometry templates are geometries defined once and reused by applying a translation, a rotation, and/or a scaling.

### 2024-05-02 paper/main.tex
streaming

@@
 However, notice that this is only true for GML files that follow the Simple Feature paradigm~\citep{OGC-SF}.
-For more complex data models like CityGML where geometry templates and \emph{XLinks} are used, streaming often needs a large amount of pre-processing.
+For more complex data models like CityGML, where geometry templates and \emph{XLinks} are used, streaming often requires a large amount of pre-processing.
 \emph{XLinks} are links between elements in a file (similar to \emph{pointers} in a computer program), a concrete example in a CityGML file is a cube that lists the geometries of 5 of its surfaces, but its 6th surface is simply a link to another surface somewhere else in the file (which belongs to another building for instance).
-If in the stream the linked surface has not be seen yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear (\eg\ to calculate its volume).
+If the linked surface has not be seen  in the stream yet, the receiver cannot process the cube and needs to wait for that specific 6th surface to appear (\eg\ to calculate its volume).
 It is not always possible to (re-)order a CityGML file so that all the references can be resolved without having to store extra information until it appears in the stream.
-However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is, in our example, copy the geometry of the 6th surface to the cube), but this means that filesize will increase, and that converting back to the original file will not be possible.
+However, it is always possible to resolve the \emph{XLinks} before streaming a dataset (that is, in our example, copy the geometry of the 6th surface to the cube), but this means that the filesize will increase, and that converting back to the original file will not be possible.
 
@@
 
-For formats that use a global indexing of vertices (as mentioned above, CityJSON is one of them, and  most formats used in computer graphics for meshes also, \eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
+For formats that use a global indexing of vertices, such as CityJSON and most formats used for storing meshes in computer graphics (\eg\ OBJ and STL), the reorganisation of the elements in a file is more complex but nonetheless possible.
 \citet{Isenburg03} describe algorithms and tools that interleave the vertices and faces in a file (instead of having one large list of vertices at the end) and add simple tags to inform that specific vertices are not used anymore in the stream (and thus can be freed from memory).

### 2024-05-02 paper/main.tex
cityjson text seq

@@
 As shown in Figure~\ref{fig:cjseq_idea}, a CityJSONSeq decomposes a CityJSON object into its features to create a sequence of several JSON objects.
-Those JSON objects are of type \texttt{CityJSONFeature}, which allows the storage of a single feature, for instance a \texttt{Building} together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}).
+Those JSON objects are of type \texttt{CityJSONFeature}, which allows the storage of a single feature, for instance a \texttt{Building}, together with its ``children'' objects (\eg\ a \texttt{BuildingPart} and/or a \texttt{BuildingInstallation}).
 Each feature is independent, it has its own list of vertices (which is thus \emph{local} to the JSON object, and is usually rather small, see next section for details) and its own textures and materials (if any).
-The allowed properties are shown in Figure~\ref{fig:cjfeature}, notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are children.
+The allowed properties are shown in Figure~\ref{fig:cjfeature}; notice that the \texttt{"id"} property is used to clearly identity the ``parent'' of the feature, in case there are children.
 \begin{figure}
@@
 \end{lstlisting}
-\caption{An example of a \texttt{CityJSONFeature} for one Building have a balcony (as its \texttt{"children"}).}%
+\caption{An example of a \texttt{CityJSONFeature} for a Building with a balcony referenced in its \texttt{"children"} property.}%
 \label{fig:cjfeature}
@@
 The rule \#4 ensures that those are available. 
-The CityJSON object must contain \texttt{"transform"} and eventually the other properties if needed; \texttt{"CityObjects"} and \texttt{"vertices"} must be present but they must be empty (to ensure that the JSON object is valid).
+The CityJSON object must contain a \texttt{"transform"} property and eventually the other properties if needed; \texttt{"CityObjects"} and \texttt{"vertices"} must be present but they must be empty (to ensure that the JSON object is valid).
 

### 2024-05-02 paper/main.tex
ch5

@@
 To convert between CityJSON and CityJSONSeq files (and vice-versa), we have developed the open-source software \emph{cjseq}, which is available at \url{https://github.com/cityjson/cjseq/} under a permissive open-source license.
-The command-line program takes care of converting not only the geometries, but also the materials, the textures, and the geometry templates that the dataset could contain.
-It contains three sub-commands: 
+The command-line program handles the conversion not only of the geometries, but also of the materials, the textures, and the geometry templates that the dataset could contain.
+It includes three sub-commands: 
 \begin{enumerate}
@@
 The files themselves are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}.
-
+% @Hugo we should find a better way to share a link to the data. 
 %
@@
 First observe that---contrary to intuition---the filesize of a dataset serialised as a CityJSONSeq file is around 12\% compacter than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\%.
-The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and do not increase to very large integers.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and they do not increase to very large integers in contrast to the vertices in CityJSON.
 For instance, the dataset \emph{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
-The fact that many indices are used for representing the geometries (and for the textures) means that if several large numbers are used then the filesize will grow; if the maximum vertex index is around 2000 for each feature then the filesize will be reduced.
+The fact that many indices are used for representing the geometries (and the textures) means that if several large numbers are used then the filesize will grow; if the maximum vertex index is around 2000 for each feature then the filesize will be reduced.
 
@@
 The scripts are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/scripts/}.
+% @Hugo I think this link is unecessary. 
 \begin{table*}
   \centering
-  \caption{Comparison of the processing time and maximum RAM use for processing CityJSON and CityJSONSeq files.
+  \caption{Comparison of the processing time and maximum RAM usage for processing CityJSON and CityJSONSeq files.
 The resident set size (RSS) is used, which is the portion of main memory occupied by the Python script.}
@@
 
-Notice that the dataset \emph{3DBV} contains not only buildings but the terrain, and a few large areas are stored as a triangulation containing a very large amount of vertices; this is the reason why the maximum memory use is larger than for other datasets.
+Notice that the dataset \emph{3DBV} contains not only buildings but also the terrain, and a few large areas are stored as a triangulation containing a very large amount of vertices; this is the reason why the maximum memory use is larger than for other datasets.
 
@@
 
-The results in Table~\ref{tab:ramtime} indicate that there is a very significant benefit to using CityJSONSeq over CityJSON, at least for operations that do not require analysing or processing city objects that are close to each other.
+The results in Table~\ref{tab:ramtime} indicate that there is a significant benefit to using CityJSONSeq over CityJSON, at least for operations that do not require analysing or processing city objects that are close to each other.
 Operations like calculating volumes, merging and subsetting files, finding city objects with specific attributes, etc.\ will all use significantly less memory and will be significantly faster.

### 2024-05-02 paper/main.tex
Fix links by adding the reproducibility repository as footnote

@@
 We have converted with \emph{cjseq} several publicly available files, and Table~\ref{tab:datasets} shows an overview of the files stored both in CityJSON and CityJSONSeq.
-The files themselves are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/data}.
-% @Hugo we should find a better way to share a link to the data. 
+The files are available in the reproducibility repository of the paper\footnote{\url{https://github.com/cityjson/paper_cjseq}}.
+
 %
@@
 The results for both the maximum memory footprint and the time used are shown in Table~\ref{tab:ramtime}.
-The scripts are available at \url{https://github.com/cityjson/paper_cjseq/tree/main/scripts/}.
-% @Hugo I think this link is unecessary. 
+The scripts are available in the reproducibility repository of the paper\footnote{\url{https://github.com/cityjson/paper_cjseq}}.
 \begin{table*}
@@
 While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for several use-cases.
+For example, CityJSONSeq is used by \emph{cjdb} (a tool for storing CityJSON datasets in a relational database) so as to store independently one feature per row in the database.
+Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\@. 
+% @Hugo: I added these examples, do we need to reference the cjdb paper?
 

### 2024-05-02 paper/main.tex
Add cjdb+3dbag papers

@@
 While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for several use-cases.
-For example, CityJSONSeq is used by \emph{cjdb} (a tool for storing CityJSON datasets in a relational database) so as to store independently one feature per row in the database.
-Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\@. 
-% @Hugo: I added these examples, do we need to reference the cjdb paper?
+For example, CityJSONSeq is used by \emph{cjdb} (a tool for storing CityJSON datasets in a relational database, see \citet{Powalka23}) to store independently one feature per row in the database.
+Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\footnote{\url{https://api.3dbag.nl/}}, which contains all buildings in the Netherlands with detailed roofs~\citep{Peters22}.
 

### 2024-05-02 paper/main.tex
Add rotterdam has high shared vertices

@@
 In CityJSON they are conceptually the same vertices and each of the surfaces refer to them, but in CityJSONSeq they have to be listed separately in each of the buildings.
-It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%, except 2 datasets have around 20\%).
+It should be said that most of the datasets have very few vertices that are shared (most have less than 2\%, except 2 datasets have around 20\%, \emph{Rotterdam} being one of them).
 To understand the correlation between the compression factor and the percentage of shared vertices in a datasets, we have modified the script to generate random cuboid buildings: the distribution of the buildings is not random, we have enforced that several buildings are adjacent to others (so that they share vertices with other buildings).

### 2024-05-02 paper/main.tex
Move part of CityJSONSeq can be dbms in the conclusion + merge with cjdb

@@
 
-Furthermore, the CityJSON specification does not prescribe the storage of CityJSONSeq, only the structure of a CityJSONSeq stream (see the five constraints above).
-In practice, CityJSONSeq can be stored in a variety of ways, eg in a single file, each feature in a separate file, in a database etc., and the optimal storage solution depends on the implementing application.
-
 %
@@
 While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for several use-cases.
-For example, CityJSONSeq is used by \emph{cjdb} (a tool for storing CityJSON datasets in a relational database, see \citet{Powalka23}) to store independently one feature per row in the database.
+
+%
+
+It should be noticed that the CityJSON specification does not prescribe the storage of CityJSONSeq, only the structure of a CityJSONSeq stream.
+In practice, CityJSONSeq can be stored in a variety of ways, for instance in a single file, each feature in a separate file, in a database, etc.
+The optimal storage solution depends on the implementing application.
+As a concrete example, we have implemented CityJSONSeq in the PostgreSQL, using \emph{cjdb}~\citep{Powalka23}, to store independently one feature per row in the database.
 Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\footnote{\url{https://api.3dbag.nl/}}, which contains all buildings in the Netherlands with detailed roofs~\citep{Peters22}.

### 2024-05-03 paper/main.tex
Just try to better format the placements of figs
City Objects: diff => misc
Make clear that Rotterdam is bigger potentially b/c 3 reasons
And make the reasons paragraphs ideas for clarity

@@
     \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
-    \textbf{3DBV}           && \qty{71634} diff    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{3DBV}           && \qty{71634} misc   &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
     \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
@@
     \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
-    \textbf{Railway}        && \qty{50} diff       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
+    \textbf{Railway}        && \qty{50} misc      & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
     \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
@@
 First observe that---contrary to intuition---the filesize of a dataset serialised as a CityJSONSeq file is around 12\% compacter than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\%.
-The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and they do not increase to very large integers in contrast to the vertices in CityJSON.
+The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and they do not increase to very large integers in contrast to the vertices in CityJSON\@.
 For instance, the dataset \emph{Helsinki} contains a total of more than 3 millions vertices, but its largest feature contains only but 2202 vertices.
@@
 Only one dataset sees its filesize slightly increase, by 4\%, when serialised to a CityJSONSeq file: \emph{Rotterdam}.
-One reason for this is that the size of the CityJSON file is rather small in the first place, and therefore the indices are not large integers.
-As an experiment, we have created around 100 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube randomly generated.
+The reasons for the increase (or decrease) are many, and we discuss in the following the 3 most relevant: (1) the total number of vertices; (2) the number of shared vertices; (3) the presence of textures.
+
+%
+
+\paragraph{Number of vertices.} 
+If a dataset has few vertices, as it is the case with \emph{Rotterdam}, then the indices will not be large integers and this might not be favourable for the compression.
+As an experiment, we have created around 100 synthetic CityJSON datasets containing buildings, and each building is represented as a simple cube, which is randomly generated.
 There are no attributes, no semantics, and no textures/materials. 
-Figure~\ref{fig:compression_random.pdfn} shows that, as the CityJSON filesize increases, the compression factor increases.
+Figure~\ref{fig:compression_random} shows that, as the CityJSON filesize increases, the compression factor increases.
 \begin{figure}
@@
 
+\paragraph{Shared vertices.} 
 The number of shared vertices between different city objects also influences the compression factor. 
@@
 If all the buildings are adjacent to another one (thus nearly 100\% of the vertices are shared), then we can see that the compression factor is about -10\% (which means that the size of the CityJSONSeq file is larger than that of the CityJSON file).
+
 %
 
+\paragraph{Textures.} 
 It should also be noticed that the attributes attached to city objects, as well as the semantics attached to surfaces, have no influence on the compression factor since they are local to each city object.
@@
 
-We compare the speed and memory footprint of accessing each city object in a dataset.
-This operation is common in applications that manipulate 3D city models.
-
-When a city model is stored in its entirety in one CityJSON object, we need to deserialise the whole CityJSON object into memory in order to access the \texttt{"transform"} and \texttt{"vertices"} properties for instance.
-
-With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one (and thus never have in memory more than the city object itself and the first JSON object in the stream).
-As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
-% Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
-
-%
-
-We have processed all the datasets from Table~\ref{tab:datasets} with two simple Python scripts that iterate through the city objects and their geometries, and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}) and report it at the end.
-This is just an example of a simple local operation, any other operation such as calculating the area of the façades or counting the number of windows could have been performed.
-The results for both the maximum memory footprint and the time used are shown in Table~\ref{tab:ramtime}.
-The scripts are available in the reproducibility repository of the paper\footnote{\url{https://github.com/cityjson/paper_cjseq}}.
 \begin{table*}
@@
 \end{table*}
+
+We compare the speed and memory footprint of accessing each city object in a dataset.
+This operation is common in applications that manipulate 3D city models.
+
+When a city model is stored in its entirety in one CityJSON object, we need to deserialise the whole CityJSON object into memory in order to access the \texttt{"transform"} and \texttt{"vertices"} properties for instance.
+
+With a CityJSONSeq file, we can read the file line by line, processing and discarding the city objects one by one (and thus never have in memory more than the city object itself and the first JSON object in the stream).
+As shown in the experiments below, this allows for very efficient operations in terms of both CPU and memory usage.
+% Operations that would highly benefit from this are, for instance, filtering and merging 3D city models.
+
+%
+
+We have processed all the datasets from Table~\ref{tab:datasets} with two simple Python scripts that iterate through the city objects and their geometries, and increment a global counter for the geometry type (\texttt{Solid} or \texttt{MultiSurface}) and report it at the end.
+This is just an example of a simple local operation, any other operation such as calculating the area of the façades or counting the number of windows could have been performed.
+The results for both the maximum memory footprint and the time used are shown in Table~\ref{tab:ramtime}.
+The scripts are available in the reproducibility repository of the paper\footnote{\url{https://github.com/cityjson/paper_cjseq}}.
 

### 2024-05-03 paper/main.tex
Improve conclusions

@@
 
-% TODO: add conclusions
-
 While CityJSONSeq was developed mostly for streaming large 3D city datasets, the fact that it has a much lower memory footprint and that it takes an order of magnitude less time to process (for some local operations) makes it an attractive alternative to CityJSON for several use-cases.
@@
 The optimal storage solution depends on the implementing application.
-As a concrete example, we have implemented CityJSONSeq in the PostgreSQL, using \emph{cjdb}~\citep{Powalka23}, to store independently one feature per row in the database.
+As a concrete example, we have implemented CityJSONSeq in the PostgreSQL database, using \emph{cjdb}~\citep{Powalka23}, to store independently one feature per row in the database.
 Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\footnote{\url{https://api.3dbag.nl/}}, which contains all buildings in the Netherlands with detailed roofs~\citep{Peters22}.
@@
 
-From a practitioners point-of-view, we should stress that CityJSONSeq can easily be processed with \emph{Unix pipes} (also called \emph{pipelines}).
+From the point-of-view of practitioners, we should stress that CityJSONSeq can easily be processed with \emph{Unix pipes} (also called \emph{pipelines}).
 Pipelines allow us to chain several processes together, the output of a process becomes the input of the next one, and so on. 
 Given 2 processes, the 2nd one can usually start before the 1st one has finished processing all the data.
-This means that, in practice, the processing of a dataset can be performed by writing several programs that perform a few small/atomic tasks; this makes the development and maintenance of code simpler.
+This means that, in practice, the processing of a dataset can be performed by writing several programs that perform a few small tasks; this makes the development and maintenance of code simpler.
 Moreover, it allows us to write the code in different languages. 
@@
 
-Finally, because the structure of CityJSONSeq is nearly the same as that of CityJSON, in practice adding support for CityJSONSeq requires a minimum amount of effort since most of the code already existing can be reused.
-We have already added support for it in a few of the open-source CityJSON tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
+Finally, because the structure of CityJSONSeq is nearly the same as that of CityJSON, in practice adding support for CityJSONSeq requires a minimum amount of effort since most of the code to parse and/or generate CityJSON objects can be reused.
+We have already added support for CityJSONSeq in a few of the open-source tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
 

### 2024-05-03 paper/main.tex
Add clean compression and thus rotate the table

@@
 \usepackage{siunitx}
+\usepackage{rotating}
 \usepackage{threeparttable}
@@
 
-\begin{table*}
+\begin{sidewaystable*}
   \centering
@@
     {@{}lcccccrrrcrrr@{}}\toprule
-    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{3}{c}{\textbf{size of file}} && \multicolumn{3}{c}{\textbf{vertices}}   \\ 
-    \cmidrule{3-4} \cmidrule{6-8} \cmidrule{10-12} 
-     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ && total & largest\footnotesize ${}^{\text{(c)}}$ & shared\footnotesize ${}^{\text{(d)}}$ \\
+    &&  \multicolumn{2}{c}{\textbf{dataset}} && \multicolumn{4}{c}{\textbf{size of file}} && \multicolumn{3}{c}{\textbf{vertices}}   \\ 
+    \cmidrule{3-4} \cmidrule{6-9} \cmidrule{11-13} 
+     && CityObjects &  app.\footnotesize ${}^{\text{(a)}}$ && CityJSON & CityJSONSeq & compr.\footnotesize ${}^{\text{(b)}}$ & clean compr.\footnotesize ${}^{\text{(c)}}$&& total & largest\footnotesize ${}^{\text{(d)}}$ & shared\footnotesize ${}^{\text{(e)}}$ \\
     \midrule
-    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
-    \textbf{3DBV}           && \qty{71634} misc   &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
-    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
-    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
-    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
-    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
-    \textbf{Railway}        && \qty{50} misc      & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
-    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\%  &&     \num{22246} &     \num{631} & 20.0\% \\
-    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
-    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
+    \textbf{3DBAG}          && \qty{1110} bldgs    &         && \qty{6.7}{\mega\byte} & \qty{5.9}{\mega\byte} & 12\% & 18\%  &&     \num{82509} &    \num{4112} &  0.1\% \\
+    \textbf{3DBV}           && \qty{71634} misc    &         && \qty{378}{\mega\byte} & \qty{317}{\mega\byte} & 16\% & 19\%  &&   \num{4110319} &  \num{116670} & 21.0\% \\
+    \textbf{Helsinki}       && \qty{77231} bldgs   &         && \qty{572}{\mega\byte} & \qty{412}{\mega\byte} & 28\% & 24\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Helsinki\_tex}  && \qty{77231} bldgs   & tex     && \qty{713}{\mega\byte} & \qty{644}{\mega\byte} & 10\% & 10\%  &&   \num{3038576} &    \num{2202} &  0.0\% \\
+    \textbf{Ingolstadt}     && \qty{55} bldgs      &         && \qty{4.8}{\mega\byte} & \qty{3.8}{\mega\byte} & 25\% & 10\%  &&     \num{87972} &   \num{12800} &  0.0\% \\
+    \textbf{Montréal}       && \qty{294} bldgs     & tex     && \qty{5.4}{\mega\byte} & \qty{4.6}{\mega\byte} & 15\% & 17\%  &&     \num{31585} &    \num{3393} &  2.0\% \\
+    \textbf{NYC}            && \qty{23777} bldgs   &         && \qty{105}{\mega\byte} &  \qty{95}{\mega\byte} & 10\% & 19\%  &&   \num{1035804} &    \num{2608} &  0.8\% \\
+    \textbf{Railway}        && \qty{50} misc       & tex+mat && \qty{4.3}{\mega\byte} & \qty{4.0}{\mega\byte} &  8\% &  9\%  &&     \num{73554} &   \num{14966} &  0.4\% \\
+    \textbf{Rotterdam}      && \qty{853} bldgs     & tex     && \qty{2.6}{\mega\byte} & \qty{2.7}{\mega\byte} & -4\% &  1\%  &&     \num{22246} &     \num{631} & 20.0\% \\
+    \textbf{Vienna}         && \qty{307} bldgs     &         && \qty{5.4}{\mega\byte} & \qty{4.8}{\mega\byte} & 11\% & 19\%  &&     \num{47220} &    \num{2025} &  0.0\% \\
+    \textbf{Zürich}         && \qty{52834} bldgs   &         && \qty{279}{\mega\byte} & \qty{247}{\mega\byte} & 11\% & 15\%  &&   \num{3472989} &    \num{4069} &  2.6\% \\
     \bottomrule
@@
       \item ${}^{\text{(b)}}$ compressi{}on factor is $\frac{size(CityJSON) - size(CityJSONSeq)}{size(CityJSON)}$
-      \item ${}^{\text{(c)}}$ number of vertices in the largest feature of the stream
-      \item ${}^{\text{(d)}}$ percentage of vertices that are used to represent different city objects
+      \item ${}^{\text{(c)}}$ clean means we remove attributes, textures, materials, and semantics.
+      \item ${}^{\text{(d)}}$ number of vertices in the largest feature of the stream
+      \item ${}^{\text{(e)}}$ percentage of vertices that are used to represent different city objects
     \end{tablenotes}
   \end{threeparttable}
-\end{table*}
+\end{sidewaystable*}
 

### 2024-05-03 paper/main.tex
Better word
footrpint

@@
 First observe that---contrary to intuition---the filesize of a dataset serialised as a CityJSONSeq file is around 12\% compacter than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\%.
+An even larger compression factor is noted in most datasets whose texture, materials, semntics and attributes have been removed. 
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and they do not increase to very large integers in contrast to the vertices in CityJSON\@.
@@
 Finally, because the structure of CityJSONSeq is nearly the same as that of CityJSON, in practice adding support for CityJSONSeq requires a minimum amount of effort since most of the code to parse and/or generate CityJSON objects can be reused.
-We have already added support for CityJSONSeq in a few of the open-source tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
+We have already added support for CityJSONSeq in a few of the open-source tools \footnote{\url{https://www.cityjson.org/software/}} and we will continue in the future.
 

### 2024-05-03 paper/main.tex
Fix Gina commit

--- 2024-05-03 paper/main.tex
@@
 Operations like modifying the CRS or updating the metadata would not even require to loop through the features, just to alter the first object in the file.
-Operations like calculating the surface of shared walls~\citep{Agugiaro22} are however not designed for streams.
+Operations like calculating the surface of shared walls~\citep{Agugiaro22} are however not suitable for streams.
 
@@
 Finally, because the structure of CityJSONSeq is nearly the same as that of CityJSON, in practice adding support for CityJSONSeq requires a minimum amount of effort since most of the code to parse and/or generate CityJSON objects can be reused.
-We have already added support for CityJSONSeq in a few of the open-source tools (see \url{https://www.cityjson.org/software/}) and we will continue in the future.
+We have already added support for CityJSONSeq in a few of the open-source tools\footnote{\url{https://www.cityjson.org/software/}} and we will continue in the future.
 

### 2024-05-03 scripts/plot_random.py
Harmonise y-axis of both fig

@@ -46,7 +46,7 @@ ax.plot(df[0], compr)
 # ax.plot(df[0], compr, marker=".")
 # ax.plot(z[:,0], compr)
 
-ax.set(xlabel='Number of buildings', ylabel='file compression')
+ax.set(xlabel='Number of buildings', ylabel='compression factor')
 # ax.xaxis.set_major_formatter(ticker.FormatStrFormatter('%.0f'))
 
 # ax.set_xlim(-1, 100000)

### 2024-05-03 paper/main.tex
Put both synthetic plots side-by-side

@@
 \documentclass{isprs} % isprs class modified 23-04-2019 (Dennis Wittich)
-\usepackage{subfigure}
+\usepackage{subcaption}
 \usepackage{setspace}
@@
 Figure~\ref{fig:compression_random} shows that, as the CityJSON filesize increases, the compression factor increases.
-\begin{figure}
+The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of more than 12\%.
+
+\begin{figure*}
   \centering
-  \includegraphics[width=\linewidth]{figs/compression_random.pdf}
-  \caption{Compression factor of CityJSONSeq files for different synthetic datasets containing only but simple cuboid buildings.}%
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=\textwidth]{figs/compression_random.pdf}
+    \caption{}%
   \label{fig:compression_random}
-\end{figure}
-The smallest file contains only 526 buildings and its compression factor is -2\% (thus CityJSONSeq has a larger filesize than that of CityJSON), while the largest file has \num{3960105} buildings, and a compression factor of more than 12\%.
+  \end{subfigure}%
+  \begin{subfigure}[b]{0.5\linewidth}
+    \centering
+    \includegraphics[width=\textwidth]{figs/compression_adjacent.pdf}
+    \caption{}%
+  \label{fig:compression_adjacent}
+  \end{subfigure}
+  \caption{Compression factor of CityJSONSeq files for different synthetic datasets.
+\textbf{(a)} Based on the number of buildings; the buildings are stored as simple cuboid that are randomly generated.
+\textbf{(b)} Based on the percentage of shared vertices.
+The same cuboid buildings are used, but we locate them adjacent to each others to create shared vertices.}%
+\label{fig:compression}
+\end{figure*}
 
@@
 The relationship between the compression and the percentage of shared vertices can be seen in Figure~\ref{fig:compression_adjacent} for around 100 datasets containing exactly \num{1000000} buildings.
-\begin{figure}
-  \centering
-  \includegraphics[width=\linewidth]{figs/compression_adjacent.pdf}
-  \caption{Compression factor of CityJSONSeq files with respect to the percentage of shared vertices in a dataset that was synthetically generated.}% 
-\label{fig:compression_adjacent}
-\end{figure}
 If the number of shared vertices is 0\% this means that we have \num{1000000} buildings that are disconnected; in this case we obtain a compression factor of around 8\% (as was the case in Figure~\ref{fig:compression_random}).

### 2024-05-03 paper/main.tex
Postgres rewrite

@@
   \caption{Compression factor of CityJSONSeq files for different synthetic datasets.
-\textbf{(a)} Based on the number of buildings; the buildings are stored as simple cuboid that are randomly generated.
-\textbf{(b)} Based on the percentage of shared vertices.
-The same cuboid buildings are used, but we locate them adjacent to each others to create shared vertices.}%
+\textbf{(a)} Based on the number of buildings; the buildings are stored as simple cuboids that are randomly generated.
+\textbf{(b)} Based on the percentage of shared vertices; the same cuboid buildings are used, but we position them adjacent to each others to create shared vertices.}%
 \label{fig:compression}
@@
 The optimal storage solution depends on the implementing application.
-As a concrete example, we have implemented CityJSONSeq in the PostgreSQL database, using \emph{cjdb}~\citep{Powalka23}, to store independently one feature per row in the database.
-Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\footnote{\url{https://api.3dbag.nl/}}, which contains all buildings in the Netherlands with detailed roofs~\citep{Peters22}.
+As a concrete example, CityJSONSeq is used by \emph{cjdb}~\citep{Powalka23}, an importer/exporter tool that stores one feature per row in PostgresSQL.
+Additionally, in order to facilitate pagination, CityJSONSeq is the return format of the 3DBAG API\footnote{\url{https://api.3dbag.nl/}}, which contains all 10 million buildings in the Netherlands with detailed roofs~\citep{Peters22}.
 

### 2024-05-06 paper/main.tex
Last typo fixed
Small improvement in abstract

@@
 
-We introduce \emph{CityJSON Text Sequences} (CityJSONSeq in short), a format based on \emph{JSON Text Sequences} and CityJSON\@. 
-CityJSONSeq was added to the CityJSON version 2.0 standard to allow us to stream very large 3D city models.
+We introduce \emph{CityJSON Text Sequences} (CityJSONSeq in short), a format based on CityJSON and \emph{JSON Text Sequences}. 
+CityJSONSeq was added to the CityJSON specifications version 2.0 to allow us to stream very large 3D city models.
 The main idea is to decompose a CityJSON dataset into its individual city objects (each building, each tree, etc.) and create several independent JSON objects of a newly defined type: \texttt{CityJSONFeature}.
@@
 First observe that---contrary to intuition---the filesize of a dataset serialised as a CityJSONSeq file is around 12\% compacter than serialised as a CityJSON file, and in the case of \emph{Helsinki} it is 28\%.
-An even larger compression factor is noted in most datasets whose texture, materials, semntics and attributes have been removed. 
+An even larger compression factor is noted in most datasets whose texture, materials, semantics and attributes have been removed. 
 The main reason for this is that the indices of the vertices are low integers for each feature (because the lowest index in each feature is always ``0'' and is incremented by 1 until the total number of vertices), and they do not increase to very large integers in contrast to the vertices in CityJSON\@.
